---
output: pdf_document #change to html_document if Latex is not installed
title: Markdown template
---

This is a Markdown template created by DIME Analytics.
It is meant to be a reference to show how text formatting in Markdown is combined with R code to produce R Markdown documents.

# Markdown

##  Headers

# This is a header
## Subheader 1
### Subheader 2
#### Subheader 3

## Paragraphs

This is a line of text.
This is another line in the same paragraph.

New paragraphs are separated by two line breaks.

## Text emphasis

Text in italics goes between *asterisks* or _underscores_.

Text in bold goes between **two asterisks** or __two underscores__.

You can combine asterisks and underscores to **emphasize with italics and bold _at the same time_**.

Strikethrough text ~~uses two tildes~~.

## Lists

Ordered lists:

1. Include a number and a dot before every item
2. Also remember to include a blank line before the beginning of the list
1. The actual number does not matter, the item will have the correct order number

Unordered lists:

* You can use an asterisk
+ Or a plus symbol
- Or a minus symbol

## Links

Include the link text in brackets followed by the URL in parentheses.
Like this:

This is [the WB website](https://www.worldbank.org) and this is the [course repository](https://osf.io/86g3b).

## Tables

Use vertical lines to separate columns and at least three dashes to separate column headers.

|This is column 1|This is column 2|
|----------------|----------------|
|Row 1           |Row 1           |
|Row 2           |Row 2           |

The width of the cells can vary in the markdown text and the output will look the same.

|This is column 1    |This is column 2|
|---|-----------|
|Row 1   |Row 1     |
|Row 2      |Row 2|

<!-- Exercise 1 -->
<!-- No code needed, just change line 2 if necessary -->

<!-- Exercise 2 -->

# R Code

```{r}
data(mtcars)
summary(mtcars$mpg)
```

<!-- Exercise 3 -->

```{r, echo = FALSE}
data(mtcars)
summary(mtcars$mpg)
```

<!-- Exercise 4 -->

# R plots

```{r, echo = FALSE}
plot(mtcars$wt,
     mtcars$mpg,
     main = "Plot example",
     xlab = "Car weight",
     ylab = "Miles per gallon")
```

<!-- Exercise 5 -->
<!-- Note that inline code doesn't go in code blocks -->

# Inline code
- The number of elements in mtcars is `r nrow(mtcars)`
- The mean of weight is `r mean(mtcars$wt)`
- The standard deviation is `r sd(mtcars$wt)`

<!-- Exercise 6 -->

# Regressions - Stargazer

```{r, echo = FALSE, message = FALSE, results = "asis"}
library(stargazer)
model <- lm(mpg ~ wt + hp, data = mtcars)
stargazer(model,
          header = FALSE,
          title = "Best table ever",
          omit = c("Constant"))
```

<!-- Exercise 7 -->

# Regressions - Huxtable

```{r, echo = FALSE, warning = FALSE}
library(huxtable)
model <- lm(mpg ~ wt + hp, data = mtcars)
huxreg(model,
       omit_coefs = c("(Intercept)"))
```
