---
title: "Coding for Reproducible Research"
subtitle: "Field Coordinator Training - R Track"
date: "June 2018"
author: "Luiza Andrade, Leonardo Viotti & Rob Marty "
output:
  beamer_presentation:
    #theme: "Pittsburgh"
    theme: "Madrid"
    colortheme: "whale"
    fonttheme: "default"
    toc: true
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F, eval=FALSE}
# Load the pacakge than contains ToothGrowth dataset
library(datasets)
```

# Introduction

## Introduction

  \begin{block}{Objective}
	  Create an R Master Script.
  \end{block} 
  
  \textbf{Content}
    \begin{enumerate}
      \item Comments
      \item Using R Studio to create a code index
      \item Indenting
      \item File paths
      \item If statements
      \item Using functions within functions
      \item Packages
      \item Loops
    \end{enumerate}

## Introduction

  \begin{itemize}
    \item The exercises on this section will help us create a Master R script
    \item To do them, go to the \textit{DataWork/Code} folder an open the file called \textit{Lab 2 - Coding for Reproducible Research.R}
  \end{itemize}

## Introduction

  \begin{itemize}
    \item Just like in Stata, a Master script must be a map of all data work
    \item It should be possible to follow all data work in the data folder, from raw data to analysis output, by following the master script
    \item It should also be possible to run all the code for the project by simply running the Master script
    \item Finally, in Stata, the Master Script is where you create globals. You can do the same in R by just creating new objects
  \end{itemize}

# Commenting

## Commenting

  \begin{itemize}
    \item Let's take a look at the script we just opened
    \item You can see that the first few lines in the script are the header, but they're not commented out
    \item In R, errors will not always break your code, so you should still be able to run this script
    \item However, not commenting out comments is still bad practice, as it makes the code harder to read
  \end{itemize}

## Commenting


   * To comment a line, write ``#`` as its first character
   * You can also add ``#`` half way through a line to comment whatever comes after it
   * In Stata, you can use ``/*`` and ``*/`` to comment part of a line's code. That is not possible in R: whatever comes after ``#`` will be a comment
   * To comment a selection of lines, press ``Ctrl + Shift + C``

## Commenting

  \begin{block}{Exercise 1}
	  Use the keyboard shortcut to comment the header of the script.
  \end{block}

  \begin{block}{Exercise 2}
	  Use the keyboard shortcut to comment the header of the script again. What happened?
  \end{block}

# Creating a document outline in RStudio

## Creating a document outline in RStudio

  \begin{itemize}
    \item RStudio also allows you to create an interactive index for your scripts
    \item To add a section to your code, create a commented line with the title of your section and add at least 4 trailing dashes, pound signs or equal signs after it
  \end{itemize}

  \begin{block}{Exercise 3}
	  Open the script index and make \texttt{PART 0} a section header. Do the same for \texttt{PART 1}.
  \end{block}

  \begin{block}{Exercise 4}
	  Note that once you create a section header, an arrow appears right next to it. Click on the arrows of parts 0 and 1 to see what happens.
  \end{block}


## Creating a document outline in RStudio

  \begin{itemize}
	  \item The outline can be accessed by clicking on the button on the top right corner of the script window. You can use it to jump from one section to another
	  \item You can also use the keyboard shortcuts \texttt{Alt + L} (\texttt{Cmd + Option + L} on Mac) and \texttt{Alt + Shift + L} to collapse and expand sections
  \end{itemize}


# Initial Settings

## Initial Settings

  \begin{itemize}
	  \item A Stata do-file typically starts with a few settings:
  \end{itemize}

```{stata, eval = F}
    clear
    set maxvar 120000
    set more off
```

## Initial Settings

  * We don't need to set the memory or the maximum number of variables in R, and ``more`` is automatically selected
  * However, if you saved the last RStudio session in ``.Rhistory``, the objects that were in RStudio's memory last time you closed it will still be there whenever you open it again
  * Therefore, it's good practice to always clean the memory when starting a new session
  * You can see all the objects currently in you memory in the *Environment* pane

## Initial Settings

  \begin{block}{Exercise 5}
    \begin{enumerate}
      \item Make sure the \textit{Environment} window is open (it should be empty now)
      \item Create an object called \texttt{foo} with any content you pick
      \item Type \texttt{ls()} to print the names of the object in memory
      \item Type \texttt{rm(foo)} to remove it
      \item To remove all objects, use \texttt{rm(list=ls())}
    \end{enumerate}
  \end{block}


# Using packages

## Using packages

  * Since there is a lot of people developing for R, it can have many different functionalities
  * To make it simpler, these functionalities are bundled into packages
  * A package is the fundamental unit of shareable code
  * It may contain new functions, but also more complex functionalities, such as a Graphic User Interface (GUI) or settings for parallel processing (similar to Stata MP)
  * They can be shared through R's official repository - CRAN (10,000+ packages reviewed and tested) and many other online sources
 * There are many other online sources such as Github, but it's important to be careful, as these probably haven't gone through a review process as rigorous as those in CRAN

## Using packages

* To install and use packages you can either do it with the user interface or by the command prompt.
```{r, eval = F}
# Installing a package
install.packages("stargazer",
                 dependencies = T)
# the dependencies argument also installs all other packages
# that it may depend upon
```

* You only have to install a package once, but you have to load it every new session. To load a package type:
```{r, eval = F}
library(stargazer)

```

## Using packages

Once a package is loaded, you can use its features and functions. Here's a list of some useful and cool packages:

* Rcmdr - Easy to use GUI
* swirl - An interactive learning environment for R and statistics.
* ggplot2 - beautiful and versatile graphics (the syntax is a pain, though)
* stargazer - awesome latex regression and summary statistics tables
* foreign - reads dtas and other formats from inferior statistical software
* zoo - time series and panel data manipulation useful functions
* data.table - some functions to deal with huge data sets
* sp and rgeos - spatial analysis
* multiwayvcov and sandwich - clustered and robust standard errors
* RODBC, RMySQL, RPostgresSQL, RSQLite - For relational databases and using SQL in R.

## Using packages

  \begin{block}{Exercise 6}
    Install the \texttt{swirl} and \texttt{stargazer} packages, including packages necessary for them to run.
  \end{block}

## Using packages

```{r, eval = F}

  # Install stargazer and swirl
  install.packages(c("stargazer", "swirl"),
                   dependencies = TRUE)

```

## Using packages

  \begin{block}{Exercise 7}
    Call the packages you just installed. Note that the \texttt{library} function only accepts one argument, so you will need to load each of them separately.
  \end{block}

## Using packages
  \scriptsize
  
```{r, eval = T}

  library(stargazer)
  library(swirl)

```

# Indenting

## Indenting

  * Indenting can be pretty different from what it looks like in Stata
  * To indent a whole line, you can select that line and press ``Tab``
  * To unindent a whole line, you can select that line and press ``Shift + Tab``
  * However, this will not always work for different parts of a code in the same line
  * In R, we typically don't introduce white space manually
  * It's rather introduced by RStudio for us

## Indenting

  \begin{block}{Exercise 8}
    To see an example of how indenting works in RStudio, add a line between the two arguments of the \texttt{install.packages} function (the vector of package names and the dependents option). Then add a line between the two package names. Note that RStudio formats the different arguments of the function differently.
  \end{block}
  
## Indenting

```{r, eval = F}
    # Same code, but this time easier to read
    install.packages(c("stargazer",
                       "swirl"),
                     dependencies = TRUE)
```

# Functions inception

## Functions inception

  * In R, you can use the output of one function as the input of another, as long as they have the same format
  * In fact, that's exactly what we just did when installing the packages
  * To see that, select just the first argument of the ``install.packages`` function and press ``Ctrl + Enter``
  * The ``c()`` function, as we know, creates a vector with its arguments


```{r, eval = T}

  c("stargazer", "swirl")

```

## Functions inception

  * The resulting vector is used as an input to the ``install.packages`` function
  * We could also have stored this vector in the memory as an object and used that object as the input
  * In fact, that's exactly what we are going to do next, so the code doesn't get too polluted as we add new packages

## Functions inception

```{r, eval = T, echo = F}
  packages <- c("stargazer",
                "swirl")

```

```{r, eval = F}

  # Create packages object
  packages <- c("stargazer",
                "swirl")

  # Use it as an input to the install.packageS() function
  install.packages(packages,
                   dependencies = TRUE)

```

# Section switches

## Section switches

  * Now, installing packages can be time-consuming, especially as the number of packages grow, and each package only needs to be installed once
  * What can we bring from our Stata Master do-files to avoid installing packages twice?

## Section switches

  * In Stata, section switches would be saved as locals
  * In R, the equivalent to that would be to create a new object

### Exercise 10
  Create a dummy scalar object called PACKAGES.
    
  *  TIP: Section switches can also be Boolean objects.

## Section switches

  * Now we need to create an if statement using this switch
  * If statements in R look like this:
```{r, eval = F}

    # Turn switch on
    PACKAGES <- 1

    # Install packages
    if (PACKAGES == 1) {
      install.packages(packages,
                       dependencies = TRUE)
    }

```

## Section switches

  * Possible variations would include
```{r, eval = F}
    # Turn switch on
    PACKAGES <- TRUE

    # Using a Boolean object
    if (PACKAGES == TRUE) {
      install.packages(packages, dep = T)
    }

    # Which is the same as
    if (PACKAGES) {
      install.packages(packages, dep = T)
    }

```

# File paths

## File paths

  * The next important part of a Master script are file paths
  * We recommend always using **explicit** and **dynamic** file paths

## File paths

  * Implicit and static file path:

```{r, eval = F}
  # Set working directory
  setwd("C:/Users/luiza/Documents/GitHub/R-Training/
        DataWork/DataSets/Final")
  
  # Load data set
  read.csv("lwh_clean.csv",
            header = T)
```

## File paths

  * Explicit and static file path:

```{r, eval = F}
  # Load data set
  read.csv("C:/Users/luiza/Documents/GitHub/R-Training/
           DataWork/DataSets/Final/lwh_clean.csv",
            header = T)
```

## File paths

  * Explicit and dynamic file path:

```{r, eval = F}
  # Define dynamic file path
  finalData <- "C:/Users/luiza/Documents/GitHub/R-Training/
                DataWork/DataSets/Final"
  
  # Load data set
  lwh <- read.csv(file.path(finalData,"lwh_clean.csv"),
                  header = T)

```
## File paths

  * File paths in R, as in Stata, are basically just strings
  * Note, however, that in R we can only use forward slashes (``/``) to separate folder names

  \begin{block}{Exercise 11}
    Let's start by adding the folder path to the training's folder in your computer to the beginning of \texttt{PART 3}.
  \end{block}

## File paths

  * You can set file paths in your master using the ``file.path()`` function
  * This function concatenates strings using ``/`` as a separator to create file paths

## File paths
  
```{r, eval = T}

  # Project folder
  projectFolder  <- 
    "C:/Users/luiza/Documents/GitHub/R-Training"

  # Data work folder
  dataWorkFolder    <- file.path(projectFolder,"DataWork")

  # Print data work folder
  dataWorkFolder

```

## File paths

  Let's check if that worked, as we will need your Master script to be running smoothly for the other sessions.

  \begin{block}{Exercise 12}
    \begin{enumerate}
      \item Turn off the switch that installs packages in your Master script
      \item Run the whole script
      \item Type the following code to open a data set using the file paths you just set:
     \end{enumerate}
  \end{block}

```{r, eval = F}
# Load data set
lwh <- read.csv(file.path(rawData,"lwh_panel.csv"),
                header = T)


```

# Looping

## Looping

  * We're almost at the end of this section
  * But the DRY rule can still be applied to part of this code

```{r, eval = F}
    # The DRY rule:
    DONT REPEAT YOURSELF

```

## Looping
  * In Stata, we'd usually use a ``foreach`` loop to go through a list of objects
  * The equivalent to that in R would be to write a for loop like this


```{r}
    # A for loop in R
    for (number in c(1.2,2.5)) {
        print(round(number))
    }
```

## Looping

  * R, however, has a whole function family that allows users to loop through an object in a more efficient way
  * They're called ``apply`` and there are many of them, for different use cases.
  * For the purpose of this training, we will only use two of them, ``sapply`` and ``apply``
  * If you look for the ``apply`` help file, you can see all of them

## Looping

  * ``sapply(X, FUN, ...)``: applies a function to all elements of a vector or list and returns the result in a vector. Its arguments are
    * **X:** a matrix (or data frame) the function will be applied to
    * **FUN:** the function you want to apply
    * **...:** possible function options

```{r}
    # A much more elegant for loop in R
    sapply(c(1.2,2.5), round)
```

## Looping

  \begin{block}{Exercise 13}
    Use the \texttt{sapply()} function to apply the \texttt{library()} function to all packages you have selected.
  \end{block}

## Looping

```{r, eval = F}
    # Load all listed packages
    sapply(packages,
           library, character.only = TRUE)
```

## Looping

A more generic version is the `apply` function.

  * ``apply(X, MARGIN, FUN, ...)``: applies a function to all columns or rows of matrix. Its arguments are
    * **X:** a matrix (or data frame) the function will be applied to
    * **MARGIN:** 1 to apply the function to all rows or 2 to apply the function to all columns
    * **FUN:** the function you want to apply
    * **...:** possible function options

## Looping


```{r, eval = T}
    # Create a matrix
    matrix <- matrix(c(1, 24, 9, 6, 9, 4, 2, 74, 2),
                       nrow = 3)

    # Look at the matrix
    matrix
```


## Looping

```{r, eval = T}
    # Row means
    apply(matrix, 1, mean)

    # Column means
    apply(matrix, 2, mean)
```

## That's all, folks

  * Now you have a template master script to use across this training's sessions
  * Save the script that you created during this session in the *DataWork* folder. Call it *MASTER.R*
  * You can run scripts from the Master script by using the ``source()`` function as you write scripts for future sessions, but we will do that on the next session

## That's all, folks

  \begin{block}{Homework}
    In the next slide here's a list of all the packages we'll need to the next sessions. Installing them might take a while, so paste them to your Master script and run it again before the next session:
  \end{block}
  
```{r, eval = T}
 packages  <- c("readstata13","foreign",
                "doBy", "broom",
                "stargazer",
                "ggplot2", "plotly", "ggrepel",
                "RColorBrewer",
                "sp", "rgdal", "rgeos", "raster", "velox",
                "ggmap", "rasterVis", "leaflet")
```


----

\huge Thank you!

----

  \frametitle{Bonus exercise}

  * You can customize your loops in R by defining your own function
  * This is done using a function conveniently called ``function()``
  * For example, if instead of just printing a number we want to print it's square, we could create a function that does both:


```{r, eval = T}
    # A much more elegant for loop in R
    sapply(c(1,2), function(x) x^2)
```

