---
title: "Data Visualization"
subtitle: "Field Coordinator Training - R Track"
date: "June 2018"
author: "Luiza Andrade, Leonardo Viotti & Rob Marty "
output:
  beamer_presentation:
    #theme: "Pittsburgh"
    theme: "Madrid"
    colortheme: "whale"
    fonttheme: "default"
    toc: true
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F, eval= T, include=FALSE}
# Packages
library(ggplot2)

if (Sys.getenv("USERNAME") == "luiza"){
  projectFolder  <- "C:/Users/luiza/Documents/GitHub/dime-r-training"
}

if (Sys.getenv("USERNAME") == "WB501238"){
  projectFolder  <- "C:/Users/WB501238/Documents/GitHub/dime-r-training"
}

if (Sys.getenv("USERNAME") == "Leonardo"){
  projectFolder  <- "C:/Users/Leonardo/Documents/GitHub/dime-r-training"

}

if (Sys.getenv("USERNAME") == "WB519128"){
  projectFolder <- file.path("C:/Users/WB519128/Documents/GitHub/dime-r-training")
}

# File paths
  dataWorkFolder    <- file.path(projectFolder,"DataWork")

  Data              <- file.path(dataWorkFolder,"DataSets")
  finalData         <- file.path(Data,"Final")
# Load CSV data
lwh <- read.csv(file.path(finalData,"lwh_clean.csv"), 
                header = T)

```


# Introduction

## Introduction

In this session, you'll learn how to use R to produce insightful, meaningful and (hopefully) nice-looking graphs. In particular, you'll use a package called ``ggplot2``.

Similarly, to the previous session, you can find some references at the end of this presentation that include a more comprehensive discussions on data visualization.

## Introduction

Before we start, let's make sure we're all set:

\begin{enumerate}

  \item Make sure you the packages \texttt{ggplot2} and \texttt{plotly} are  installed and loaded.
  
  \begin{itemize}
    \item Preferably, do this by adding them to \texttt{packages} vector of your Master script.
  \end{itemize}
 
  \item Load the \texttt{lwh\_clean.csv} data set that were created in the last session.
    \begin{itemize}
      \item Run the Master script to load all the necessary packages and set file paths.
      \item Remeber to disable the \texttt{PACKAGES} switch in the Master if you already installed them. This will save you a lot of time.
    \end{itemize}
  
\end{enumerate}

## Introduction
\scriptsize

```{r, eval= F}
  # Install packages
  install.packages(c("ggplot2","plotly"),
                   dependencies = TRUE)
  
  # Load packages
  library(ggplot2)
  library(plotly)
  
  # Load CSV data
  projectFolder  <- "YOUR/FOLDER/PATH"
  finalData <- file.path(projectFolder,
                         "DataWork","DataSets","Final")
  lwh <- read.csv(file.path(finalData,"lwh_clean.csv"),
                  header = TRUE)
  lwh_simp <- read.csv(file.path(finalData,"lwh_simp.csv"),
                       header = TRUE)

```

## Introduction

In our workflow there are usually two distinct uses for plots:

  \begin{enumerate}
    \item \textbf{Exploratory analysis:} Quickly visualize your data in an insightful way
      \begin{itemize}
        \item We'll do this using R's base plot, that allows you to quickly create some basic plots
      \end{itemize}
    \item \textbf{Publication/Reporting:} Make pretty graphs for a presentation, a project report, or to just show your boss something other than the laziest graph you could come up with
      \begin{itemize}
        \item We'll do this using \texttt{ggplot}, a flexible and powerful tool to plot beautiful graphs
        \item \texttt{ggplot}'s syntax is more complicated, but it's easier to make your graphs look good with it
      \end{itemize}
  \end{enumerate}


# Exploratory graphs - Base plot

## Exploratory graphs - Base plot

  For exploratory plots, we're going to use Base plot. It is easy to use and can produce useful graphs with very few lines of code.

## Exploratory graphs - Base plot

### Exercise 1:
Plot the ``lwh_simp`` data set you constructed in the previous session using the ``plot()`` function.

 1. If you don't have the code from the previous session. Here it is:

```{r, eval = T}

covariates <- c("age_hhh",
                "num_dependents",
                "income_total_trim",
                "expend_food_yearly",
                "w_area_plots_a")

lwh_simp <- lwh[,covariates]
```
 2. Now, pass the name of the data set as the only argument for the plot function

## Exploratory graphs - Base plot

\scriptsize

```{r}
plot(lwh_simp)
```

## Exploratory graphs - Base plot

\begin{block}{Exercise 2}
  \begin{enumerate}
    \item Create a histogram of variable \texttt{w\_area\_plots\_a} in data frame \texttt{lwh\_simp}. Use the \texttt{hist()} function for that.
    \item Try to set the color as \texttt{"green"}. Use the help if you are not sure how to do it.
  \end{enumerate}
\end{block}


## Exploratory graphs - Base plot
\scriptsize
```{r, eval = T, echo = T}
hist(lwh_simp$w_area_plots_a, col = "green")
```

## Exploratory graphs - Base plot

### Exercise 3
Use the ``boxplot()`` function to see a description of the yearly food expenditure by number of dependents.

- TIP: The syntax for this function is ``variable_to_be_plotted`` ~ ``variable_to_plot_over``.

## Exploratory graphs - Base plot
\scriptsize

```{r}
boxplot(lwh_simp$expend_food_yearly ~ lwh_simp$num_dependents)

```

# ``ggplot`` - Introduction

## ``ggplot`` - Introduction


The ``ggplot2`` package is an implementation of the theoretical framework for the construction of quantitative graphs developed in The Grammar of Graphics, a book by Leland Wilkinson. 

It is a powerful and easy to use tool (once you understand its logic) that produces complex and multifaceted plots.

## ``ggplot`` - Introduction

There are a few reasons why we use ggplot:

 * It is generally easier to customize and tailor plots to your needs.
 * It works great with almost any kind of plot, including maps, with little variation in syntax. The notable exception is 3D plotting, which can be done, but it's not as straightforward.
 * It looks good.

## ``ggplot`` - Grammar

Here is a list of ``ggplot2`` most commonly used "grammar" elements:


\begin{table}[]
\centering
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|m{2cm}| m{8cm}|}
  \hline
  \rowcolor{myColor}
  \multicolumn{1}{|c|}{\textcolor{white}{Element}}   &      \multicolumn{1}{|c|}{\textcolor{white}{Description}} \\
  Data       & The dataset(s) being used.  \\ \hline
  Aesthetics & How your data is maped. For example what goes in the X and Y axis, size, shape and color. \\ \hline
  Geometries & The visual elements used to represent the data (e.g. lines, dots, bars, etc.) \\ \hline
  Themes     & All non-data formating. \\
  \hline
  \end{tabular}
\end{table}



## ``ggplot`` - Data structure

As almost everything in R, ``ggplot`` is very flexible and can be used in different ways. It is easier to work, however, if your data is in a specific format.

  * Data has to be a data frame.

  * Long format, especially categorical values.
    + It's better to have one variable with all the categories than a set of dummies.
  * Labelled factors.

This is very similar to Stata, with the advantage of not needing to
preserve and restore anything.

##``ggplot`` - Data structure

Here's an example of how an aggregated (collapsed) data set should look like:


```{r, include = T, echo = F, warning=FALSE}
anualIncGen <- aggregate(x= lwh["income_total_trim"],
                         by = list(year = lwh$year,
                                   gender =  lwh$gender_hhh),
                         FUN = mean, na.rm = T)
anualIncGen

```


## ``ggplot`` - Introduction


### Exercise 4

First, let's create a very simple plot with the yearly expenditure on food on the Y axis and age of the household head on the X axis.


 1. Use the ``ggplot`` function to store your plot in an object called ``plot1``
    - Since it's the first one, Here's the code:

```{r, eval=T}
plot1 <- ggplot(data = lwh,
                aes(y = w_area_plots_b,
                    x = w_gross_yield_b))
```

 2. Now, try to print your plot. What happens?
    - You can do it by using the ``print()`` function or just typing the object name in the console

## ``ggplot`` - Introduction

We mapped our variables, but without a geometry ``ggplot`` doesn't know how to represent the data.

### Let's add a geometry
```{r, eval=FALSE}
ggplot(data = lwh,
       aes(y = w_area_plots_b,
           x = w_gross_yield_b)) +
  geom_point()
```


## ``ggplot`` - Introduction
\scriptsize

```{r, echo = F, warning = F}
ggplot(lwh, aes(y = w_area_plots_b,
                x = w_gross_yield_b)) +
  geom_point()
```

# ``ggplot`` - Data structure

## ``ggplot`` and ``aggregate()``

* Since ``ggplot`` usually works better with data in a specific format, there are a few functions particularly useful to manipulate data
* Here, we'll use the ``aggregate()`` function
* ``aggregate()`` works similarly to ``collapse`` in Stata

## ``ggplot`` and ``aggregate()``

To create a line plot with ``ggplot``, we will create a new data set containing the average of the total house hold annual income by year.
```{r, eval= T}
# Aggregate anual income by year
anualInc <-
    aggregate(x   = lwh["income_total_trim"], # data.frame
              by  = list(year = lwh$year), #list
              FUN = mean, na.rm = T) # function
```

### Exercise 5
 1. Use the ``ggplot()`` function with the ``anualInc`` as the ``data`` argument, ``income_total_trim`` as ``y`` and ``year`` as ``x`` in ``aes()``
 2. This time add ``geom_line()``


## ``ggplot`` and ``aggregate()``
\scriptsize

```{r,  eval= F}

# Plot code with arguments in geom
ggplot() +
    geom_line(data = anualInc, 
              aes(y = income_total_trim,
                  x = year))


```

## ``ggplot`` and ``aggregate()``
\scriptsize

```{r, echo = F , eval= T}

ggplot() +
    geom_line(data = anualInc, 
              aes(y = income_total_trim,
                  x = year))



```

## ``ggplot`` and ``aggregate()``

Ok, that worked, but was a really boring plot. Let's do a cooler one.

### Exercise 6: Part 1

Use the ``aggregate()`` function to average total house hold annual income, now, by two variables: `year` and `treatment_hh`.

- TIP: add an argument to the `list()` function in the by argument. Like this:

```{r, eval = F}
  by = list(Year = lwh$year, 
            Treatment = lwh$treatment_hh)
```

## ``ggplot`` and ``aggregate()``
\scriptsize

```{r}
# Aggregate data set
anualIncGen <- aggregate(x= lwh["income_total_win"],
                         by = list(year = lwh$year,
                                   treatment =  lwh$treatment_hh),
                         FUN = mean, na.rm = T)

anualIncGen

```

## ``ggplot`` and ``aggregate()``

### Exercise 6: Part 2
 Now, use `ggplot` as in the previous exercise to plot the aggregated data frame.
 
 1. This time set the ``color`` and `group` arguments in the ``aes()`` function as the treatment variable.
 2. Finally, add both line and point ``geoms``.

## ``ggplot`` and ``aggregate()``
\scriptsize

```{r,  eval= F}
# Aggregate data set
anualIncGen <- aggregate(x= lwh["income_total_win"],
                         by = list(year = lwh$year,
                                   treatment =  lwh$treatment_hh),
                         FUN = mean, na.rm = T)

# Plot aggregated data set
ggplot(anualIncGen, 
       aes(y = income_total_win,
           x = year,
           color = treatment,
           group = treatment)) +
  geom_line() +
  geom_point()

```

## ``ggplot`` and ``aggregate()``
\scriptsize

```{r, echo = F , eval= T, warning=FALSE}

anualIncGen <- aggregate(x= lwh["income_total_win"],
                         by = list(year = lwh$year,
                                   treatment =  lwh$treatment_hh),
                         FUN = mean, na.rm = T)

ggplot(anualIncGen, aes(y = income_total_win, x = year, color = treatment, group = treatment)) +
  geom_line() +
  geom_point()


```

# ``ggplot`` - Aesthetics

## ``ggplot`` - Aesthetics

In ``ggplot`` , the aesthetic element defines how data is represented in the aesthetics (e.g. color, shape and size) of geometric objects (e.g. points, lines and bars).

### Exercise 7: Part 1

 1. Create a data frame containing only 2018 observations of ``lwh`` in the Rwamangana 35 site.

  - TIP: This time we want to keep only observations (lines) that meet a certain condition. Use the left argument of the brackets operators. Like this: 
```{r, eval = F}
    lwh[CONDITION,]
```

 2. Use ``ggplot()`` to scatter plot area in season A and total annual income

## ``ggplot`` - Aesthetics
\scriptsize

```{r,  eval= F}
# Subset lwh to 2018 and Rwamangana 35
lwh_s <- lwh[lwh$year == 2018 & lwh$site_code == "Rwamangana 35", ]

# Plot
ggplot(lwh_s, aes(x = w_area_plots_a,
                  y = income_total_trim)) +
    geom_point()

```

## ``ggplot`` - Aesthetics
\scriptsize

```{r, echo = F , eval= T , warning=FALSE}
# Subset lwh to 2018 and Rwamangana 35
lwh_s <- lwh[lwh$year == 2018 & (lwh$site_code == "Rwamangana 35"), ]

# Plot
ggplot(lwh_s, aes(x = w_area_plots_a, y = income_total_trim)) +
    geom_point()

```


## ``ggplot`` - Aesthetics

Now add the women's dietary diversity score[^1] to ``aes()``.


### Exercise 7: Part 2

 1. Use the ``wdds_score`` varaible as color in ``aes()``.
    - TIP: this is is a categorical variable, but it is stored in numeric format.  Use the ``factor()`` function. Like this:
```{r, echo = T , eval= F}
    color = factor(wdds_score)
```

[^1]:
Women's dietary diversity score is based on the questions asked specifically to an adult female respondent about her food consumption. We ask if she ate 16 food items categorized into 9 groups. Each group get 1 if any of the food belonging to the group is consumed. The final score is a simply sum of the categories.

## ``ggplot`` - Aesthetics
\scriptsize

```{r,  eval= F}
# Plot code
ggplot(data = lwh_s,
       aes(y = w_area_plots_a,
           x = income_total_trim,
           color = factor(wdds_score))) +
    geom_point()

```

## ``ggplot`` - Aesthetics
\scriptsize

```{r, echo = F , eval= T, warning=FALSE}

ggplot(data = lwh_s,
       aes(y = w_area_plots_a, x = income_total_trim, color = factor(wdds_score))) +
    geom_point()

```

## ``ggplot`` - Aesthetics

Now we will combine different arguments of ``aes()``

### Exercise 7: Part 3

 1. Add expenditure on food last week variable, ``expend_food_lastweek`` to the ``size`` argument of ``aes()`` in your last graph.

## ``ggplot`` - Aesthetics
\scriptsize

```{r,  eval= F}
# Plot code
ggplot(data = lwh_s,
       aes(y = w_area_plots_a,
           x = income_total_trim,
           color = factor(wdds_score),
           size = expend_food_lastweek)) +
    geom_point()
```

## ``ggplot`` - Aesthetics
\scriptsize

```{r, echo = F , eval= T, warning=FALSE}

ggplot(lwh_s, aes(y = w_area_plots_a, x = income_total_trim, color = factor(wdds_score), size = expend_food_lastweek)) +
    geom_point()

```

## ``ggplot`` - Aesthetics

* Because we used a factor variable as the color in the aesthetic, it displayed different categories clearly
* If we had used a numeric variable instead, it would have created a gradient
* You'll not create this for time's sake, but here's how it would look:

```{r,  eval= F}
# Plot code
ggplot(data = lwh_s,
       aes(y = w_area_plots_a,
           x = income_total_trim,
           color = w_gross_yield_a,
           size = expend_food_lastweek)) +
    geom_point()
```

## ``ggplot`` - Aesthetics
\scriptsize

```{r, echo = F , eval= T, warning=FALSE}

ggplot(lwh_s, aes(y = w_area_plots_a, x = income_total_trim, color = w_gross_yield_a, size = expend_food_lastweek)) +
    geom_point()

```

# ``ggplot`` - Titles and labels

## ``ggplot`` - Titles and labels

Lastly, we can add titles, axis labels and format legends. This is done by adding additional layers to the plot.

To add a title and legend to the X and Y axis, we use the functions listed below. They all take a string scalar as argument.

 * ``ggtitle()`` - adds a title to the plot
 * ``xlab()`` - adds a label to X axis
 * ``ylab()`` - adds a label to Y axis

## ``ggplot`` - Titles and labels

To add legend titles, we will use two functions that control the formatting of the aesthetic. As they can take many different arguments, we use ``name`` argument sets legend title:

 * ``scale_color_discrete() `` - formatting of the color aesthetic.
 * ``scale_size_continuous()`` - formatting of the color aesthetic.

## ``ggplot`` - Titles and labels

#### Exercise 8

 1. Copy the code for the graph produced in the previous exercise.
 2. Use the ``+`` symbol to add the layer.
 3. ``ggtitle()``, ``xlab()`` and ``ylab()`` take simple strings as inputs.
 4. For ``scale_color_continuous()`` and ``scale_size_continuous()`` you need to specify ``name`` argument as the desired string.

## ``ggplot`` - Titles and labels
\scriptsize

```{r, eval= F}
# A properly labelled plot
ggplot(lwh_s, aes(y = w_area_plots_a,
                  x = income_total_trim,
                  size  = expend_food_lastweek,
                  color = factor(wdds_score))) +
  geom_point() +
  ggtitle("My pretty plot") +
  xlab("Total Income (Wins.)") +
  ylab("Plot area in season A (Wins.)") +
  scale_color_discrete(name = "Women's dietary diversity score") +
  scale_size_continuous(name = "Food exp. last week")
```

## ``ggplot`` - Titles and labels
\scriptsize

```{r, echo = F , eval= T,  warning=FALSE}
ggplot(data = lwh_s,
       aes(y = w_area_plots_a,
           x = income_total_trim,
           color = factor(wdds_score),
           size = expend_food_lastweek)) +
  geom_point() +
  ggtitle("My pretty plot") +
  xlab("Total Income (Wins.)") +
  ylab("Plot area in season A (Wins.)") +
  scale_color_discrete(name = "Women's dietary diversity score") +
  scale_size_continuous(name = "Food exp. last week")
```

# Saving a plot

## Saving a plot

Let's save the previous plot using the PDF graphics device

#### Exercise 9:
  1. Open the PDF graphics device by using the ``pdf()`` function.
  2. Set the ``file`` argument as your outputs path with the file name (including ".pdf")
  3. Paste the code of the previous plot bellow the ``pdf()`` function.
  4. Finally, use the ``dev.off()`` to close the device, i.e. tell R that you're finished plotting.

## Saving a plot
\scriptsize

```{r, eval= F}
  # Open PDF graphics device
  pdf(file = file.path(rawOutput, "plot1.pdf"))

   # Plot
   ggplot(data = lwh_s,
       aes(y = w_area_plots_a,
           x = income_total_trim,
           color = factor(wdds_score),
           size = expend_food_lastweek)) +
     geom_point() +
     ggtitle("My pretty plot") +
     xlab("Total Income (Wins.)") +
     ylab("Plot area in season A (Wins.)") +
     scale_color_discrete(name = "Women's dietary diversity score") +
     scale_size_continuous(name = "Food exp. last week")

  # Close PDF graphics device
  dev.off()
```


# Interactive graph

## Interactive graph

Now we'll use ggplotly() to create an interactive graph!

### Bonus exercise - ggplotly
  1. Load the ``plotly`` package
  2. Store one the plots produced in the previous ``ggplot`` exercises in an object.
  3. Pass that object as argument for the ``ggplotly()`` function

## Interactive graph
\scriptsize

```{r,  eval= F}
# Load package
library(plotly)

# Store graph in plot1
plot1 <- ggplot(data = lwh_s,
                aes(y = w_area_plots_a,
                    x = income_total_trim,
                    color = factor(wdds_score))) +
    geom_point()

# Use ggplotly to create an interactive plot
ggplotly(plot1)
```

# References and recommendations

## References and recommendations

Websites:

  * The R Graph Gallery:
  https://www.r-graph-gallery.com/

  * Gpplot official site:
  http://ggplot2.tidyverse.org/


Online courses:

  * Johns Hopkins Exploratory Data Analysis at Coursera:
  https://www.coursera.org/learn/exploratory-data-analysis

Books:

  * The grammar of graphics by Leland Wilkinson.

  * Beautiful Evidence by Edward Tufte.

  * R Graphics cook book by Winston Chang

  * R for Data Science by Hadley Wickham andGarrett Grolemund

# Appendix

## Appendix

\large Slides that didn't make the cut

## R common graphic devices

R has several built-in graphic devices. A graphic device is the medium where you visually represent a plot. Here are the most come:

  * The standard screen device or the plot pane in RStudio.
  * PDF file (file device).
  * PNG (file device).
  * JPEG (file device).

## Differed ways of saving a plot

There are a couple of ways of saving plots in R:

 * With ggplot you can use the ``ggsave()`` function that has a simple syntax and can do most of the dirty work for you.

 * Choosing an specific graphics device (other than the default window). pdf(), png(), jpeg() and bmp() are functions that call graphic devices that save your plots in these formats. There are a few others built-in and you can check them by typing *?Devices*, and even more in CRAN.


## Graphic systems

R has four main graphic systems:

  * Base plot - most basic plotting system.
  * Grid graphics - is a low-level graphics system mainly used by package developers.
  * Lattice graphics - easy way to plot high dimensional data or many graphs at the same time.
  * Ggplot2 - a flexible and powerful tool to plot beautiful graphs with intuitive syntax

