---
title: "Descriptive analysis"
subtitle: "R for Stata Users"
date: "November-December 2018"
author: "Luiza Andrade, Leonardo Viotti & Rob Marty "
output:
  beamer_presentation:
    #theme: "Pittsburgh"
    theme: "Madrid"
    colortheme: "whale"
    fonttheme: "default"
    #toc: true
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F, eval=T, include=FALSE}
library(stargazer)
library(tidyverse)

# File paths

if (Sys.getenv("USERNAME") == "luiza"){
  projectFolder  <- "C:/Users/luiza/Documents/GitHub/dime-r-training"
  
}

if (Sys.getenv("USERNAME") == "WB501238"){
  projectFolder  <- "C:/Users/WB501238/Documents/GitHub/dime-r-training"
  
}

if (Sys.getenv("USERNAME") == "Leonardo"){
  projectFolder  <- "C:/Users/Leonardo/Documents/GitHub/dime-r-training"

}

if (Sys.getenv("USERNAME") == "WB519128"){
  projectFolder <- file.path("C:/Users/WB519128/Documents/GitHub/dime-r-training")
}

  # File paths
  dataWorkFolder    <- file.path(projectFolder,"DataWork")

  Data              <- file.path(dataWorkFolder,"DataSets")
  finalData         <- file.path(Data,"Final")
  rawOutput         <- file.path(dataWorkFolder,"Output","Raw")

  # Load CSV data
  whr <- read.csv(file.path(finalData,"whr_panel.csv"), 
                header = T)
  
```

# Introduction
  
## Introduction
  
Descriptive statistics are used to represent the basic features of data. When we talk about descriptive analysis, it usually means that we're not making any assumptions, and we're not using probability theory to infer anything beyond the immediate data.

This session is mostly focused on how to implement descriptive analysis in R. We will not go in depth into these concepts, but you can find some useful references at the end of this presentation.

## Introduction

This session will cover two topics:
  
  \begin{enumerate}
    \item Quick ways to extract summary information from your data
    \item How to use this information to create and export tables
  \end{enumerate}
  

# Getting started

## Getting started
First, let's load the data that is going to be used in the training.

### Load the data
```{r, include = T, results = "hide", eval = F}
  #### Load CSV data

  # Replace with where your data is
  FolderPath <- file.path("YOUR FOLDER PATH HERE")

  whr <- read.csv(file.path(FolderPath,"whr_panel.csv"), 
                  header = T)

```

## Getting started

Before starting, lets install the packages we'll use in this session since it might take a while.

We'll get into what packages are in a moment.

### Install today's packages

```{r, include = T, results = "hide", eval = F}
  install.packages(c("stargazer", 
                     "tidyverse",
                     "openxlsx"), 
                   dependencies = T)

```

# Using packages

## Using packages

  * Since there is a lot of people developing for R, it can have many different functionalities.
  * To make it simpler, these functionalities are bundled into packages.
  * A package is just a unit of shareable code.


## Using packages

  * It may contain new functions, but also more complex functionalities, such as a Graphic User Interface (GUI) or settings for parallel processing (similar to Stata MP).
  * They can be shared through R's official repository - CRAN (13,000+ packages reviewed and tested).
 * There are many other online sources such as GitHub, but it's important to be careful, as these probably haven't gone through a review process as rigorous as those in CRAN.

## Using packages


* To install and use packages you can either do it with the user interface or by the command prompt.
```{r, eval = F}
# Installing a package
install.packages("stargazer",
                 dependencies = T)
# the dependencies argument also installs all other packages
# that it may depend upon to run
```


* You only have to install a package once, but you have to load it every new session. 


## Using packages

Once a package is loaded, you can use its features and functions. Here's a list of some useful and cool packages:

* swirl - An interactive learning environment for R and statistics.
* tidyverse - a collection of R packages designed for data science.
    + ggplot2 - beautiful and versatile graphics. The syntax is a bit of pain, but it can fufill all of graphics dreams.
* Rcmdr - Easy to use GUI. 
* stargazer - awesome latex regression and summary statistics tables.
* foreign - reads dtas and other formats from inferior statistical software.
* sp, sf, rgeos, raster, rgdal - spatial analysis.
* RODBC, RMySQL, RPostgresSQL, RSQLite - For relational databases and using SQL in R.


## Using packages

### Exercise 1
 1. Now load the packages we just installed. Use the `library()` function to do it.
    + Note that the `library` function only accepts one argument, so you will need to load each of them separately.

```{r, eval = T}

  library(stargazer)
  library(tidyverse)
  library(openxlsx)

```

## Warinings vs Errors

What if this happens?

\begin{figure}
  \includegraphics[scale=0.5]{img/Warning.png}
\end{figure}


## Warinings vs Errors

 
R has two types of error messages, `warnings` and actual `erros`:

  * `Errors` - break your code, i.e., prevent it from running.
  * `Warnings` - usually mean that nothing went wrong yet, but you should be careful. 
  
RStudio's default is to print warning messages, but not stop the code at the lines where they occur. You can configure R to stop at warnings if you want.  
  

# Quick summary statistics


## Quick summary statistics


`summary(x, digits)` - equivalent to Stata's *summarize*, displays summary statistics. Its arguments are:

 * **x:** the object you want to summarize, usually a vector or data frame
 * **digits:** the number of decimal digits to be displayed
    
\begin{block}{Exercise 1}
  Use the \texttt{summary()} function to display summary statistics for the \textit{whr} data frame.
\end{block}

## Quick summary statistics

\scriptsize

```{r, eval = F}
  # Summary statistics
  summary(whr)
```

```{r, eval = T, echo = F}
  # Summary statistics
  summary(whr[,1:7])
```
 
## Quick summary statistics
 
 
``table()`` - equivalent to `tabulate` in Stata, creates a frequency table. Its main arguments are the objects to be tabulated.

\begin{block}{Exercise 2}
  Use the \texttt{table()} function to display frequency tables for:
    \begin{enumerate}
      \item The variable \textit{year} in the \textit{whr} data frame
      \item The variables \textit{region} and \textit{year} in the \textit{whr} data frame, simultaneously
    \end{enumerate}
\end{block}


## Quick summary statistics
```{r, eval = T}
  # Year of data collection
  table(whr$year)
```

## Quick summary statistics
\scriptsize


```{r, eval = T}
  # Gender of household head per year
  table(whr$region, whr$year)
```


# Descriptives tables

## Descriptives tables


 We can also use the \texttt{stargazer()} function to quickly display a nice-looking descriptives table.

  \textit{Stargazer} was originally developed to export beautiful regression tables to \LaTeX\ or html, but it also allows you to generate summary statistics.

 
## Descriptives tables

\begin{block}{Exercise 3 - \texttt{stargazer()} summary statistics table}
  Use the \texttt{stargazer()} function  to display summary statistics for the variables in the \textit{whr} data frame.
\end{block}

  The \texttt{stargazer()} function accepts \textbf{a lot} of arguments, most of which are beyond the scope of this session. Here are the arguments you'll need for this specific table:
  \begin{itemize}
    \item \textbf{x:} the object you want to summarize -- in this case a vector or data frame
    \item \textbf{type:} the output format -- "text" to just display, "latex" (the default) to save as a \LaTeX\ table, and "html" for, yes, html
    \item \textbf{digits:} the number of decimal digits to be displayed
  \end{itemize}

## Descriptives tables
\tiny

```{r, eval = T}
# A descriptive table with stargazer
stargazer(whr, 
          digits = 1, 
          type = "text")

```



```{r, include = T, echo = F}
# stargazer(whr[1:15], 
#           digits = 1, 
#           type = "text")

```



# Export tables to \LaTeX


## Export tables to \LaTeX\


To export the table to \LaTeX\, we will use a couple of additional arguments of the `stargazer()` function:

  \begin{itemize}
    \item \textbf{out:} where to save the table, i.e., the file path, including the file name
    \item \textbf{covariate.labels:} a vector of variable labels
  \end{itemize}

## Export tables to \LaTeX\

But first, let's pick a few variables of interest in the `whr` data set so the table fits in these slides.

### Exercise 2

 1. Create a vector called ``covariates`` containing the string names of the variables you want to keep: `happy_score`, `gdp_pc`, `family`, and `trust_gov_corr`.
 2. Use this vector to subset the whr dataset to contain only these variables. Call the new data frame `whr_simp`.


## Export tables to \LaTeX\
\scriptsize


```{r}
# Vector with covariates to be kept
covariates <- c("happy_score",
                "gdp_pc",
                "freedom",
                "trust_gov_corr")

# subset whr
whr_simp <- whr[, covariates]

```


## Export tables to \LaTeX\

### Exercise 3: Now use the stargazer function to export the `whr_simp`:

 1. Create a vector `cov_labels` containing labels for the `happy_score`, `gdp_pc`, `freedom` and `trust_gov_corr` variables.
 2. Set ``whr_simp`` as the `x` argument this time
 3. Set the `covariate.labels` argument as the vector you just created

## Export tables to \LaTeX\

\tiny

``` {r, eval = T, results = "hide"}

  # Set labels
  cov_labels <- c("Happy score", "GDP per capita",
                  "Freedom", "Trust in gornment and currption")

  # Save table to latex
  stargazer(whr_simp,
            # Formatting
            covariate.labels = cov_labels,
            summary.stat = c("n", "mean", "sd", "min", "max"), # Chose which stats to comopute
            digits = 2,
            # You can directly export with the out argument
            out = file.path(rawOutput,"desc_table.tex"))

```

\input{"../DataWork/Output/Raw/desc_table.tex"}

# Descriptives tables - Create tables from scratch

## Descriptives tables - Create tables from scratch

In R, it is relatively easy to construct any table you can think of by manipulating objects. To construct a table from scratch, we will use two functions:

 * `aggregate()` - Similar to collapse in Stata, it can compute statistics of a variable based on the values of other variable.
 * `spread()` - Reshapes data sets from long to wide. [^1] 

[^1]: `gather()` Reshapes data sets from wide to long, both are from `tidyverse`. There are other ways to reshape data, but these are becoming the standard in R.

## Descriptives tables - Create tables from scratch

`aggregate(X, by, FUN)`:

 * __x__: a data frame or column
 * __by__: a list of grouping variables
 * __FUN__: a function to compute statistics

## Descriptives tables - Create tables from scratch

### Exercise 3
 Use the `aggregate()` function to create a data frame called `happy_table` with the mean of `happy_score` per year and region. 

```{r, eval = T, results = "hide"}
# Aggregate income by year and treatment status
happy_table <-
  aggregate(happy_score ~ year + region,
            data = whr,
            FUN = mean)
```


## Descriptives tables - Create tables from scratch
\scriptsize

```{r, eval = F}
print(happy_table)

```


```{r, echo= F}
print(happy_table[1:19,])

```

## Descriptives tables - Create tables from scratch

`spread(data, key, value)`:
 
 * __data__: a data frame
 * __key__: the variables that identify the group in the wide data set
 * __value__: the variable in long format that has multiple records from the same group or individual

## Descriptives tables - Create tables from scratch

### Exercise 4
 Use the `spread` function to make the `happy_table` data frame wide in the `year` variable.


```{r, eval = T, results = "hide"}
# Aggregate income by year and treatment status
happy_table <- 
  spread(happy_table, 
         key = year, 
         value = happy_score)
```

For comparison, here's how you'd do it in Stata:
\textcolor{red}{CHECCK}

``reshape wide x, i(year) j(region)``


## Descriptives tables - Create tables from scratch
\scriptsize

```{r}
print(happy_table)

```

## Descriptives tables - Create tables from scratch


With a data frame as input, `stargazer` by default tries to summarize it. So, to export this table we must specify one additional argument: `summary = F`.

### Exercise 4
Print the ``happy_table`` table you created in exercise 6 using ``stargazer``. If you want, you can also save it using the ``out`` option.


## Descriptives tables - Create tables from scratch
\scriptsize

```{r, eval = F}

# Create table
stargazer(happy_table,
          summary = F,
          # Exporting
          out = file.path(...),
          # Formatting:
          title = "Happy table",
          digits = 1,
          rownames = F)
```


## Descriptives tables - Create tables from scratch
\scriptsize

```{r, echo = F, results='asis'}

# Create table
stargazer(happy_table,
          summary = F,
          header = F,
          title = "Happy table",
          digits = 1,
          rownames = F)
```


# Export tables to Excel

## Export tables to Excel


There are several ways to export R objects to Excel. We will use here the the `write.xslx()` function of the `openxlsx` package.

 It takes a matrix or data frame object as input and saves it as a `.xlsx` file.

`write.xslx()` is one of the most common functions, but there are many other functions that allow you to export formatted tables to Microsoft Excel, Word or PowerPoint. Here are some examples:

  * ReporteRs
  * Flextable
  * r2excel (only available in GitHub).

## Export tables to Excel

  `write.xlsx(x, file, row.names = TRUE, col.names ...)` 
  
 * __x__:	the object to be written
 * __file__: where to save the table, i.e., the file path including the file name
 * __row.names__: a logical value indicating whether the row names of x are to be written along with x

## Export tables to Excel

### Exercise 5

Use the `write.xlsx()` function to save the `happy_table` you table created in Exercise 4 into an xlsx file.

1. Set `x` argument as *happy_table*.
2. Set `file` as the folder path to your output folder including a name for the file plus ".xlsx"

 Tips:
 
 * Use the help function to check syntax if needed

## Export tables to Excel
\scriptsize

```{r, include = T, eval = F}
write.xlsx(happy_table,
           file = file.path(rawOutput, "happy_table.xlsx"))

```


\begin{figure}[H]
  \centering
  \includegraphics[scale=.7]{img/happy_table_xlsx.png}
\end{figure}


# Export regression table

## Export regression table

\begin{block}{Warning:}
This is a session on \textbf{descriptive} analysis, so regressions are beyond its scope.
\end{block}

But since you'll probably ask, here's how you run a regression and how you export a very simple regression table to \LaTeX\ using `stargazer` and the `iris`[^2] dataset:

```{r}
# Run a Regression
reg1 <- lm(Sepal.Length ~ Sepal.Width,
           data = iris)

```

[^2]: `iris` is a built in data set in R with some morphology information for the flowers of 3 iris species.

## Export regression table
\scriptsize


```{r, eval = F}
# Export a regression table

depvar_label <- "Sepal Length"
covar_labels <- "Sepal Width"

stargazer(reg1,
          title = "Regression table",
          dep.var.labels = depvar_label,
          covariate.labels = covar_labels,
          digits = 2,
          # out = "..."
          header = F)

```



## Export regression table
\tiny

```{r, results ='asis', echo = F}
# Export a regression table
depvar_label <- "Sepal Length"
covar_labels <- "Sepal Width"

stargazer(reg1,
          title = "Regression table",
          dep.var.labels = depvar_label,
          covariate.labels = covar_labels,
          digits = 2,
          header = F)

```



## Export regression table
\scriptsize

```{r}

# Regression 1
reg1 <- lm(Sepal.Length ~ Sepal.Width,
           data = iris)

# Reg with year FE
reg2 <- lm(Sepal.Length ~ Sepal.Width + Petal.Length,
           data = iris)

# Reg with year and region FE
reg3 <- lm(Sepal.Length ~ Sepal.Width + Petal.Length + factor(Species),
           data = iris)

```

## Export regression table
\scriptsize

```{r, eval = F}
depvar_label <- "Sepal Length"
covar_labels <- c("Sepal Width",
                  "Petal Length")
#Table
stargazer(reg1,
          reg2,
          reg3,
          font.size = "tiny",
          title = "Regression table",
          keep = c("Sepal.Width",  "Petal.Length"),
          dep.var.labels = depvar_label,
          covariate.labels = covar_labels,
          add.lines = list(c("Species FE", "No", "No", "Yes")),
          omit.stat = c("ser"),
          digits = 2,
          header = F)
```

## Export regression table
\scriptsize

```{r, results ='asis', echo = F}
depvar_label <- "Sepal Length"
covar_labels <- c("Sepal Width",
                  "Petal Length")
#Table
stargazer(reg1,
          reg2,
          reg3,
          font.size = "tiny",
          title = "Regression table",
          keep = c("Sepal.Width",  "Petal.Length"),
          dep.var.labels = depvar_label,
          covariate.labels = covar_labels,
          add.lines = list(c("Species FE", "No", "No", "Yes")),
          omit.stat = c("ser"),
          digits = 2,
          header = F)
```

# References and recommendations

## References and recommendations


  * Johns Hopkins Exploratory Data Analysis at Coursera:
  https://www.coursera.org/learn/exploratory-data-analysis

  * Udacity's Data Analysis with R:
  https://www.udacity.com/course/data-analysis-with-r--ud651

  * Jake Russ stargazer cheat sheet:
  https://www.jakeruss.com/cheatsheets/stargazer/

## References and recommendations

  Since we talked about \LaTeX\ so much...

  * DIME \LaTeX\ templates and trainings:
  https://github.com/worldbank/DIME-LaTeX-Templates

  * All you need to know about \LaTeX\:
  https://en.wikibooks.org/wiki/LaTeX
  
  
# Appendix

## Appendix - Comments and shortcuts

 * There is only one characterto comment out code in R, `#`.
 * It can be used both by itself or after a command:
```{r}
# This is a comment
foo <- 42 # This is also a comment

```
 
 * `Ctrl + Shift + c` is a shortcut to comment out a larger chunk of code. Simmilar to `/*` and `*/` in Stata.
 
 * `Ctrl + Enter` is a shortcut to run the currently select code.

## Appendix - Save your data to .csv

\textcolor{red}{update}

`write.csv(x, file = "", sep = ",", row.names = TRUE)`

 * __x__:	the object to be written
 * __file__: where to save the table, i.e., the file path including the file name
 * __sep__: the field separator of the csv, Excel's default is comma
 * __row.names__: either a logical value indicating whether the row names of x are to be written along with x, or a character vector of row names to be written


## Appendix - Save your data to .csv
```{r, eval = F}

  write.csv(whr,
            file = file.path(...,"whr.csv"),
            row.names = F)

```

## Appendix - Formulas

Formulas are a way of describing a relationship between variables or objects. They work as inputs for several functions, notably regression functions.


### We can create formulas by using the formula function
```{r}
# or Formula function yield same results
formula1 <- formula(y ~ x1 + x2)
formula1
```



## Appendix - Formulas

The mont basic structure of a formula is actually just the tilde symbol ~ and at least one right-hand variable.

### You can also covert strings to create formulas
```{r}
# or Formula function yield same results
formula2 <- as.formula("~ x1")
formula2
```

## Appendix - Formulas

Note that values that assigned to the symbols in the formula are not accessed when the formula is created.

Alternatively, if you write an expression containing a tilde R already understands it as a formula.

### Just using the tilde
```{r}

formula3 <- y ~ x1 + z1
formula3
```


