---
title: "Introduction I - R basics"
subtitle: "R for Advanced Stata Users"
author: "DIME Analytics"
date: "The World Bank | [WB Github](https://github.com/worldbank) <br> `r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: ["libs/remark-css/default.css", "libs/remark-css/metropolis.css", "libs/remark-css/metropolis-fonts.css"]
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=4, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T#, echo=F, warning=F, message=F
  )
library(tidyverse)
library(hrbrthemes)
library(fontawesome)
library(xaringanExtra)
xaringanExtra::use_panelset()
```

```{r xaringan-logo, echo=FALSE}
xaringanExtra::use_logo(
  image_url = "img/lightbulb.png",
  exclude_class = c("inverse", "hide_logo"),
  width = "50px"
)
```

```{r, echo = F, eval=T}
# Load the package than contains ToothGrowth dataset
library(datasets)

# File paths

if (Sys.getenv("USERNAME") == "WB532468"){
  projectFolder  <- "C:/WBG/Repos/dime-r-training"
  
}

if (Sys.getenv("USERNAME") == "WB501238"){
  projectFolder  <- "C:/Users/WB501238/Documents/GitHub/dime-r-training"
  
}

if (Sys.getenv("USERNAME") == "Leonardo"){
  projectFolder  <- "C:/Users/Leonardo/Documents/GitHub/dime-r-training"

}

if (Sys.getenv("USERNAME") == "wb519128"){
  projectFolder <- file.path("C:/Users/WB519128/Documents/GitHub/dime-r-training")
}

# File paths
  dataWorkFolder    <- file.path(projectFolder,"DataWork")

  Data              <- file.path(dataWorkFolder,"DataSets")
  finalData         <- file.path(Data,"Final")
# Load CSV data
# lwh <- read.csv(file.path(finalData,"lwh_clean.csv"), 
#                 header = T)

whr <- read.csv(file.path(finalData,"whr_panel.csv"), 
                header = T)

```

# Table of contents

1. [Installation](#installation)

2. [Dummy Slides]

---
name: installation

# Installation

## Installation

This training requires that you have R installed in your computer:

### Instructions

 * Please visit (https://cran.r-project.org) and select a Comprehensive R Archive Network (CRAN) mirror close to you.

 * If you're in the US, you can directly visit the mirror at Berkley university at (https://cran.cnr.berkeley.edu).

 * we also strongly suggest installing R studio. You can get it in (https://www.rstudio.com/), but you need to install R first.