---
title: "Data Processing"
subtitle: "R Training for NISR"
date: "Aug. 2018"
author: "Luiza Andrade, Leonardo Viotti & Rob Marty "
output:
  beamer_presentation:
    #theme: "Pittsburgh"
    theme: "Madrid"
    colortheme: "whale"
    fonttheme: "default"
    toc: true
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include=FALSE}
# Run Master
GitHub  <- "~/GitHub/dime-r-training"
source(file.path(GitHub, "Presentations/presentations_master.R"))
```

# Introduction

## Introduction

* In this session, you'll be introduced to some basic concepts of data cleaning in R. The contents covered are:
    * Importing data
    * Exploring a data set
    * Uniquely and fully indentifiable ID variables
    * Creating new variables
    * Dropping variables
    * Subsetting a data set
    * Dealing with factor variables
    * Aggregating variables
    * Treating outliers
* There are many other tasks that we usually perform as part of data cleaning that are beyond the scope of this session

## Introduction

Before we start, let's make sure we're all set:

\begin{enumerate}
  \item Make sure the packages \texttt{readstata13} and \texttt{dplyr} are listed in the \texttt{packages} vector of your Master script. If they are not installed (they should be!), install them.
  \item If you haven't yet in this session, run the Master script to load all the necessary packages and set file paths.
  \item Remeber to disable the \texttt{PACKAGES} switch in the Master if you already installed them. This will save you a lot of time.
  \item Open the script called \texttt{Lab 3} in NISR R Training/Codes. We've provided you with some code to build upon that will save you some time during this session.
\end{enumerate}

## Introduction

Here's a shortcut if you missed the last session:

\scriptsize

```{r, eval= F}
  # Install packages
  install.packages(c("readstata13","dplyr"),
                   dependencies = TRUE)

  # Load packages
  library(dplyr)
  library(readstata13)

  #  Set folder paths
  trainingFolder  <- "YOUR/FOLDER/PATH"
  finalData       <- file.path(trainingFolder, "Data", "Final")
  rawlData        <- file.path(projectFolder, "Data", "Raw")

```

# Loading a data set

## Loading a data set from CSV

* In R, we usually save data in CSV format
* The .rds format is also a nice alternative, which we will cover soon, but it isn't the standard
* CSV files can be a lot lighter than binary files
* It is also possible to do version control of CSV files in .git
* On the other hand, the data they store is in a much simpler format, and we'll see some  shortcomings of that soon
* To load a CSV data set, we use the ``read.csv()`` function

## Loading a data set from CSV

### ``read.csv(file)``
 
 * **file**: is the path to the file you want to open, including its name and format (``.csv``)
 
### Exercise 1
Load the ``lwh_panel.csv`` data set from "NISR R Training/Data/Raw". Create an object called ``panel`` with this data set.

 * TIP: use the ``file.path()`` function and the ``rawData`` object created in the master to simplify the folder path.
         

## Loading a data set from CSV

```{r}
  # Load a CSV data set
  panel <- read.csv(file.path(rawData, "lwh_panel.csv"))
```

If you look at the ``Environment`` pane, you'll be able to see this data set. Click on the blue arrow to see the list of  variables.


\begin{figure}
\centering
  \includegraphics[scale=0.4]{img/panel.png}
\end{figure}


## Loading a data set from CSV

* Note that R reads string variables as factors as default
* This format saves memory, but can be tricky if you actually want to use the variables as strings
* You can specify the option ``stringsAsFactors = FALSE`` to prevent R from turning strings into factors
* You can also simply recast any variables you want to be strings (as they'll probably be just a few) using the ``as.character()`` function

## Loading a Stata data set

* You can also load a ``.dta`` file, i.e., a Stata data set, using the ``read.dta13()`` function.
* This function takes exactly the same argument as the ``read.csv`` function

### Exercise 2
Use the function ``read.dta13()`` to load the ``endline_data_raw.dta`` data set from "NISR R Training/Data/Raw". Create an object called ``endline`` with this data set.


## Loading a Stata data set
\scriptsize

```{r}
  # Load the raw data from LWH endline
 endline <- read.dta13(file.path(rawData,
                                 "endline_data_raw.dta"))
```

## Loading a Stata data set

* In Stata, you can have different levels of a labelled variable with the same value label. That's not true for R factors, which is why all those warnings were created
* If you go to the ``Environment`` pane and click on the blue arrow, you'll see that string variables were imported as strings, and only labelled values were imported as factors
* If you scroll down on the variables list, you can also find some metadata such as variable labels, value labels and Stata formats saved as variables' attributes

# Exploring a data set

## Exploring a data set

Some useful functions:

* **``class()``:** reports object type or type of data stored
* **``dim()``:** reports the size of each one of an object's dimension
* **``names()``:** returns the variable names of a data set
* **``str()``:** general information on a R object, similar to ``codebook`` in Stata

## Exploring a data set

### Exercise 3
Use some of the functions listed in the previous slides to explor the ``endline`` and ``panel`` objects.

* TIP: all of thesefunctions take a single argument, which is the object to be described.

## Exploring a data set
\scriptsize

```{r}
  # See objects' formats
  class(endline)
  class(panel)

  # How many observations and variables?
  dim(endline)
  dim(panel)
  
```

## Exploring a data set
\scriptsize
```{r, eval = F}
  # Variable names
  names(endline)
```

```{r, eval = T, echo = F}
  # Variable names
  names(endline)[1:12]
```

```{r, eval = F}
  str(panel)
```
```{r, eval = T, echo = F}
  # Data set structure
  str(panel[, 1:5])
  
  # Variable structure
  str(panel$treatment_hh)
```

# ID variables

## ID variables

Desired properties of an ID variable: *uniquely and fully identifying*

* An ID variable cannot have duplicates
* An ID variable may never be missing
* The ID variable must be constant across a project
* The ID variable must be anonymous

## ID variables

Some useful functions to identify an ID variable in R

* **unique():** displays unique occurrences in an object
* **duplicated():** returns a boolean vector showing which observations of the object are duplicated
* **length():** length of an object
* **is.na():** returns a boolean vector showing which observations of the object have missing values

## ID variables

### Exercise 4
Use one of the functions in the previous slide to tell if ``hh_code`` is a uniquely and fully identifying variables for the ``panel`` data set.

* TIP: All these functions take a single argument, which is the object we're testing.

## ID variables

  * Uniquely identifiying?
  
  \scriptsize
```{r, eval = F}
  # See all unique values of hh_code
  duplicated(panel$hh_code)
```

```{r, eval = T, echo = F}
  # See all unique values of hh_code
  duplicated(panel$hh_code)[1:20]
```

```{r}
  # Count duplicates
  sum(duplicated(panel$hh_code))
```

## ID variables

  * Fully identifiying?
  
  \scriptsize
```{r, eval = F}
  # See all unique values of hh_code
  is.na(panel$hh_code)
```

```{r, eval = T, echo = F}
  # See all unique values of hh_code
  is.na(panel$hh_code)[1:20]
```

```{r}
  # Count NAs
  sum(is.na(panel$hh_code))
```


## ID variables

Ok, that was a tricky question, as we already knew this is a panel data set. So let's try with what is a more credible ID variable

### Exercise 5
Count the number of missing observations and duplicates values in the variables ``hh_code`` and ``year`` in the ``panel`` data set.

* TIP: you can use ``panel[,c("hh_code","year")]`` as the function's argument to test both variables at the same time. Note that this is still a single argument, but this time it's a data frame with two columns.

## ID variables

\scriptsize

```{r}
# Uniquely identifying?
sum(duplicated(panel[,c("hh_code","year")]))
# Fully identifiying?
sum(is.na(panel$year))
sum(is.na(panel[,c("hh_code","year")]))
```

## ID variables

### Exercise 6
Create new variable called ``id`` in the ``panel`` data set containing a uniquely and fully identifiable ID. 

  * TIP: To create a new variable in a data set, we simply assign a value to a column that doesn't year exist. You can make the ID unique by concatenating the year and the household ID.

## ID variables

\scriptsize 
```{r}
  # Create panel ID
  panel$id <- (panel$hh_code * 10000) + panel$year
  
  # Check properties
  sum(duplicated(panel$id))
  sum(is.na(panel$id))

  # Here's a shortcut similar to the isid variable in Stata
  length(unique(panel$id, na.rm = TRUE)) == length(panel$id)
```

# Subsetting

## Subsetting

* The ``panel`` has a lot more variables than we'll actually use for analysis
* Let's construct a new data set with only the variables we'll you
* For starters, we'll subset the ``panel`` data set and keep only some variables of interest
* These variables are already listed in the script we gave you
* But first let's take a look at how to find variables to select in a data set

## Subsetting

The ``grep`` function is similar to the wildcard (``*``) in Stata: it looks for all elements of an object containing a certain pattern of strings.

``grep(pattern, x)``: finds the elements of ``x`` that contain the pattern in ``pattern`` and returns a vector of indexes with their position

* **`pattern`:** the string expression you want to look for
* **`x`:** a vector of strings

## Subsetting

### Exercise 7
Create a vector called ``plot_area_vars`` containing the names of the variables with winsorized plot areas in seasons A, B and C:

1. Use the ``grep()`` function to display the position index of all the variables in the ``panel`` data set whose name contains the expression ``area``. 

2. Use the resulting vector from the ``grep()`` function to subset the vector of all variable names in the ``panel`` data frame.

3. Create the ``plot_area_vars`` vector containing only the names of the winsorized plot area variables

* TIP: these variables' names start with "w_".

## Subsetting
```{r, eval =F}
# Get plot area variables indexes
grep("area", names(panel))

# Get plot area variables names
names(panel)[grep("area", names(panel))]

# These are the ones we want:
plot_area_vars <- c("w_area_plots_a",
                    "w_area_plots_b",
                    "w_area_plots_c")

```


## Subsetting

\scriptsize

* In the previous sessions, we talked about indexing and how we can use it to select specific rows and columns of objects
* Here's a quick recall:

```{r, eval = F}
  ## Select a column (variable) in a data frame:
  # First column
  panel[, 1]

  # Select by name
  panel[, "hh_code"]
  panel$hh_code
  
  ## Select elements in a vector
  
  # Select one element in a vector
  panel$hh_code[2]
  
  # Select multiple elements in a vector
  panel$hh_code[c(3,6,89)]
  
  # Which is the same as
  panel[c(3,6,89), "hh_code"]
```

## Subsetting

* Now, let's use this information to subset our ``panel`` data frame and keep only the variables we have already listed

### Exercise 8
Create a new object, called ``lwh``, containing only the variables we selected from the ``panel`` data set.


## Subsetting
\tiny
```{r}
  # Select variables
  id_vars          <- c("hh_code", "wave", "year", "treatment_hh",
                        "treatment_site", "site_code")
  demographic_vars <- c("gender_hhh", "age_hhh", "num_dependents", "read_and_write")
  yield_vars       <- c("w_gross_yield_a", "w_gross_yield_b")
  food_vars        <- c("expend_food_yearly", "expend_food_lastweek", "wdds_score")
  income_vars      <- names(panel)[28:37]
  plot_area_vars   <- c("w_area_plots_a", "w_area_plots_b", "w_area_plots_c")

 # Subset data
 lwh <- panel[, c(id_vars,
                  demographic_vars,
                  yield_vars,
                  food_vars,
                  income_vars,
                  plot_area_vars)]

  # Check result
  names(lwh)
```

# Factor variables

## Creating a factor

Let's take a look at the variable ``gender_hhh`` in the ``lwh`` data set:

```{r}
  # Tabulate gender of household head
  table(panel$gender_hhh)
```

This is supposed to be categorical variable, but for now it's just numeric.

## Creating a factor

To create a factor variable, we use the ``factor`` function:

``factor(x, levels, labels)`` : turns numeric or string vector ``x`` into a factor vector 
* **``x``:** the vector you want to turn into a factor
* **``levels``:** a vector containing the possible values of ``x``
* **``labels``:** a vector of strings containing the labels you want to apply to your factor variable

### Exercise 9
Replace the variable ``gender_hhh`` in the ``lwh`` data set by a factor. Label the factor so that 0 is "Female" and 1 is "Male".

* TIP: the first string in the ``labels`` argument will be applied to the first value in the ``levels`` argument, the second to the second and so on.


## Creating a factor
\scriptsize
```{r}
  # The numeric variable
  table(lwh$gender_hhh)

  # Turn numeric variable into factor
  lwh$gender_hhh <- factor(lwh$gender_hhh,
                           levels = c(0, 1),
                           labels = c("Female", "Male"))
  
  # The factor variable
  table(lwh$gender_hhh)
```

## Exploring factors

Here are a few useful functions to explore factors:

 * **``str()``:** we've already seen this function. When using it with factors is that is shows the number and some variable labels
 * **``levels()``:** displays the different values of a factor variable
 * **``table()``:** displays a frequency table per level of a factor variable

### Exercise 10

1. Use the functions listed above to explore the ``wave`` variable in the ``lwh`` data set.
2. (challenge) Use the ``table()`` function to see the year of each wave. Use the help if necessary.

## Exploring factors
\tiny
```{r}
  # First look at the variable
  str(lwh$save)

  # List all levels
  levels(lwh$wave)
  
  # Observations per level
  table(lwh$wave)
  
  # Year of each wave
  table(lwh$wave, lwh$year)
```

## Ordering factors

* The ``wave`` variable's levels are not ordered correctly
* That's because R creates the number underlying a factor in alphabetical order
* However, that can be misleading, specially if we're using this variable for tables and graphs
* If we want the levels of a factor to be in an specific order, we need to set the argument ``ordered`` of the ``factor()`` function to ``TRUE``
* This will tell R that the order of the levels is exactly the same as the one specified in the ``levels`` argument


## Ordering factors

### Exercise 11
Make the ``wave`` variable in the ``lwh`` dataset an ordered factor, ordering them by the year in which they happened

## Ordering factors

\scriptsize

```{r}
  # Order variable
  panel$wave <- factor(panel$wave,
                       levels = c("Baseline", "FUP1&2", "FUP3", "FUP4", "Endline"),
                       ordered = T)
  
  # Check new order
  levels(panel$wave)
  table(panel$wave, panel$year)
```


# Aggregating variables

## Aggregate variables

* For our analysis in the next sessions, we will not use the disaggregated income variables
* We'll rather be interested in the total income of a household
* To calculate the total income, we'll use the ``rowSums()`` function, equivalent to ``rowtotal`` in Stata
* ``rowSums()`` is part of a function family that allows you to aggregate row and columns of a data set
* They all have the same syntax

## Aggregate variables

``rowSums(x, na.rm = FALSE)``: sums selected variables in all rows of a data set

* **``x``:** the two-dimensional object whose columns you want to add
* **``na.rm``:** by default, when any of the values you're trying to add are missing, the result will be a missing value. Make this argument ``TRUE`` for R to treat missing values as zero in the sum


## Aggregate variables

### Exercise 12
Create a variable called ``income_total`` in the ``lwh`` data set containing the sum of all the income variables you listed in the ``income_vars`` vector.

## Aggregate variables

\scriptsize

### Exercise 13
Create a variable called ``income_total`` in the ``lwh`` data set containing the sum of all the income variables you listed in the ``income_vars`` vector.

```{r}
# List variables
income_vars

# Add columns
lwh$income_total <-
  rowSums(lwh[, income_vars],
          na.rm = TRUE)
```

# Dropping Variables

## Dropping variables

Now, let's remove the disaggregated income variables from data set, since we're using them anymore.

The easiest way to remove variables from a data set is to make those variables ``NULL``. Like this:

```{r, eval = F}
dataset$variable <- NULL
```

To do this for a series of variables at the same time, you can select them by their names. Like this:
```{r, eval = F}
dataset[, c("var1", "var2", "var3")] <- NULL
```

## Dropping variables

### Exercise 14
Remove all of the raw income variables listed in the ``income_vars`` vector from the ``lwh`` data data set.

## Dropping variables

### Exercise 15
Remove all of the raw income variables listed in the ``income_vars`` vector from the ``lwh`` data data set.

```{r}
  # Remove income variables
  lwh[, income_vars] <- NULL
```

# Treating outliers

## Treating outliers

* There are different ways to treat outliers and there's not a single best method for doing it
* In the end, this is an analysis question, it is good to avoid droping a whole observation from your data set because of an outlier in a single variable

## Winsorizing

To winsorize a variable means to replace all observations with value above (or below) a given percentile by the value of that percentile, therefore capping the distribution of the resulting variable.

There are a few different functions that perform winsorization in R, but we'll do it here by writing our own.

As we've seen in session 2, you can create functions in R using a function called ``function()``. In the next slide, we'll give you some code to create a function that

1. Calculates the value of the 90th percentile of its argument
2. Replaces all values above this percentile in its argument by the 90th percentile
3. Returns the resulting variable


## Winsorizing

```{r}
 # Create a function to winsorize at the 90th percentile
 winsor <- function(x) {
    x[x > quantile(x, 0.9, na.rm = T)] <-
      quantile(x, 0.9, na.rm = T)
    return(x)
  }

  # Create winsorized income
  lwh$income_total_win <- winsor(lwh$income_total)
```

## Trimming

To trim a variable means to replace all observation above a certain value (usually a given percentile) with missing values. It's particularly useful to create non-distorted graphs.

### Exercise 16
Edit the ``winsor()`` function created earlier to create a new function that trims observations:

1. Instead of replacing the values with the 90th quantile, replace it with "NA"
2. Call this function ``trim``
3. Use this function to trim the ``income_total`` variable in the ``lwh`` data set and create a new variable called ``income_total_trim``

## Trimming

```{r}

  # Create the trim function
  trim <- function(x) {
    x[x > quantile(x, 0.9, na.rm = T)] <- NA
    return(x)
  }

  # Trim income
  lwh$income_total_trim <- trim(lwh$income_total)

```

## Treating outliers

```{r}
  # Compare variables
  summary(lwh$income_total)
  summary(lwh$income_total_win)
  summary(lwh$income_total_trim)
```

# Saving a data set

## Saving a data set as CSV

* As mentioned before, R data sets are usually save as CSVs
* To save a data set, we use the ``write.csv()`` function:

### ``write.csv(x, file, row.names = TRUE)``
* **``x``:** the object (usually a data frame) you want to export to CSV
* **``file``:** the file path to where you want to save it, including the file name and the format (".csv")
* **``row.names``**: by default, R adds a column to the CSV file with the names (or numbers) of the rows in the data frame. Set it to ``FALSE`` if you don't want that column to be exported

## Saving a data set as CSV

### Exercise 17
Save the ``lwh`` data set to NISR R Training/Data/Final.

* TIP: Use the ``file.path()`` function and the object ``finalData`` created in the master to simplify the folder path.

## Saving a data set as CSV

```{r, eval = F}
# Save the lwh data set
write.csv(lwh,
          file.path(finalData,"lwh_clean.csv"),
          row.names = F)
```

## Saving a data set in R format

* The main issue, at least for now, of saving a data set as CSV is that when you open it again all strings and factors will be imported in the same format (either strings or factors, depending on how you set the ``stringsAsFactors`` option). 
* One other issue is that factors will be labelled in alphabetical order, so any ordering of factors that you may have created will be lost 

```{r, eval = T}
# Save the lwh data set as .RDS
save(lwh, 
     file = file.path(finalData, "lwh_clean.rds"))
```

# Running a script from the master

## Running a script from the master

### Exercise 18

1. Save this script
2. Add a switch for Lab3 in your master
3. Create an if statement that runs this script if the switch is on. To do this, use the ``source()`` function, which is equivalent to the ``do`` function in Stata.

* TIP: just like the ``do`` function in Stata, the only argumento of the ``source()`` function is the complete file path to your script. Don't forget to you explicit and dynamic file paths!

