---
title: "Descriptive analysis"
subtitle: "Field Coordinator Training - R Track"
date: "June 2018"
author: "Luiza Andrade, Leonardo Viotti & Rob Marty "
output:
  beamer_presentation:
    #theme: "Pittsburgh"
    theme: "Madrid"
    colortheme: "whale"
    fonttheme: "default"
    toc: true
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F, eval=T, include=FALSE}
library(stargazer)

if (Sys.getenv("USERNAME") == "luiza"){
  projectFolder  <- "C:/Users/luiza/Documents/GitHub/dime-r-training"
}

if (Sys.getenv("USERNAME") == "WB501238"){
  projectFolder  <- "C:/Users/WB501238/Documents/GitHub/dime-r-training"
}

if (Sys.getenv("USERNAME") == "Leonardo"){
  projectFolder  <- "C:/Users/Leonardo/Documents/GitHub/dime-r-training"

}

if (Sys.getenv("USERNAME") == "WB519128"){
  projectFolder <- file.path("C:/Users/WB519128/Documents/GitHub/dime-r-training")
}

  # File paths
  dataWorkFolder    <- file.path(projectFolder,"DataWork")

  Data              <- file.path(dataWorkFolder,"DataSets")
  finalData         <- file.path(Data,"Final")
  rawOutput         <- file.path(dataWorkFolder,"Output","Raw")

  # Load CSV data
  lwh <- read.csv(file.path(finalData,"lwh_clean.csv"), 
                  header = T)
```

# Introduction
  
## Introduction
  
Descriptive statistics are used to represent the basic features of data. When we talk about descriptive analysis, it usually means that we're not making any assumptions, and we're not using probability theory to infer anything beyond the immediate data.

This session is mostly focused on how to implement descriptive analysis in R. We will not go in depth into these concepts, but you can find some useful references at the end of this presentation.

## Introduction

This session will cover two topics:
  
  \begin{enumerate}
    \item Quick ways to extract summary information from your data
    \item How to use this information to create and export tables
  \end{enumerate}
  

## Introduction

First, let's load the data that is going to be used in the training. Paths should be set in your master file!

#### Load the data
```{r, include = T, results = "hide", eval = F}
  # Load CSV data
  lwh <- read.csv(file.path(finalData,"lwh_clean.csv"), 
                  header = T)

```

# Quick summary statistics


## Quick summary statistics


`summary(x, digits)` - equivalent to Stata's *summarize*, displays summary statistics. Its arguments are:

 * **x:** the object you want to summarize, usually a vector or data frame
 * **digits:** the number of decimal digits to be displayed
    
\begin{block}{Exercise 1}
  Use the \texttt{summary()} function to display summary statistics for the \textit{lwh} data frame.
\end{block}

## Quick summary statistics

\scriptsize

```{r, eval = F}
  # Summary statistics
  summary(lwh)
```

```{r, eval = T, echo = F}
  # Summary statistics
  summary(lwh[,1:8])
```
 
## Quick summary statistics
 
 
``table()`` - equivalent to `tabulate` in Stata, creates a frequency table. Its main arguments are the objects to be tabulated.

\begin{block}{Exercise 2}
  Use the \texttt{table()} function to display frequency tables for:
    \begin{enumerate}
      \item The variable \textit{year} in the \textit{lwh} data frame
      \item The variables \textit{gender\_hhh} and \textit{year} in the \textit{lwh} data frame, simultaneously
    \end{enumerate}
\end{block}


## Quick summary statistics
```{r, eval = T}
  # Year of data collection
  table(lwh$year)
```

## Quick summary statistics
```{r, eval = T}
  # Gender of household head per year
  table(lwh$gender_hhh, lwh$year)
```


# Descriptives tables

## Descriptives tables


 We can also use the \texttt{stargazer()} function to quickly display a nice-looking descriptives table.

  \textit{Stargazer} was originally developed to export beautiful regression tables to \LaTeX\ or html, but it also allows you to generate summary statistics.

 It can also be used to export any data frames you create to \LaTeX\ as a formatted table. To do that, you fist need to construct a data frame object combining vectors (of the same length) with the desired information.
 
## Descriptives tables

 
### If you haven't yet done this in your master, install and load the ``stargazer`` package now!
 
```{r, eval = F}

  # Install stargazer
  install.packages("stargazer",
                   dependencies = TRUE)

  # Load stargazer
  library(stargazer)
```


## Descriptives tables

\begin{block}{Exercise 3 - \texttt{stargazer()} summary statistics table}
  Use the \texttt{stargazer()} function  to display summary statistics for the variables in the \textit{lwh} data frame.
\end{block}

  The \texttt{stargazer()} function accepts \textbf{a lot} of arguments, most of which are beyond the scope of this session. Here are the arguments you'll need for this specific table:
  \begin{itemize}
    \item \textbf{x:} the object you want to summarize -- in this case a vector or data frame
    \item \textbf{type:} the output format -- "text" to just display, "latex" (the default) to save as a \LaTeX\ table, and "html" for, yes, html
    \item \textbf{digits:} the number of decimal digits to be displayed
  \end{itemize}

## Descriptives tables
\tiny

```{r, eval = F}
# A descriptive table with stargazer
stargazer(lwh, 
          digits = 1, 
          type = "text")

```



```{r, include = T, echo = F}
stargazer(lwh[1:15], 
          digits = 1, 
          type = "text")

```



# Export tables to \LaTeX

## Export tables to \LaTeX\


To export the table to \LaTeX\, we will use a couple of additional arguments of the `stargazer()` function:

  \begin{itemize}
    \item \textbf{out:} where to save the table, i.e., the file path, including the file name
    \item \textbf{covariate.labels:} a vector of variable labels
  \end{itemize}

## Export tables to \LaTeX\

But first, let's pick a few variables of interest in the `lwh` data set so the table fits in these slides.

### Exercise 4
 1. Create a vector called ``covariates`` containing the string names of the variables you want to keep: `age_hhh`, `num_dependents`, `income_total_win`, and `expend_food_yearly`.
 2. Use this vector to subset the lwh dataset to contain only these variables. Call the new data frame `lwh_simp`.


## Export tables to \LaTeX\
\scriptsize

```{r}
# Vector with covariates to be kept
covariates <- c("age_hhh",
                "num_dependents",
                "income_total_win",
                "expend_food_yearly")

# subset lwh
lwh_simp <- lwh[, covariates]

```


## Export tables to \LaTeX\

### Exercise 4
 1. Create a vector called ``cov_labels`` containing the labels to the covariates, in the same order as in the ``covariates`` vector.

 2. Now use the stargazer function as in the previous exercise:
  + Set ``lwh_simp`` as the `x` argument this time
  + Set the `covariate.labels` argument as the vector you just created

## Export tables to \LaTeX\

\tiny

``` {r, eval = T, results = "hide"}

  # Set labels
  cov_labels <- c("Age of household head", "Number of dependents",
                  "Anual income (winsorized)", "Yearly food expediture")

  # Save table to latex
  stargazer(lwh_simp,
            covariate.labels = cov_labels,
            #summary.stat = c("n", "mean", "sd", "min", "max"),
            digits = 1,
            out = file.path(rawOutput,"desc_table.tex"))

```

\input{"../DataWork/Output/Raw/desc_table.tex"}

# Descriptives tables - Create tables from scratch

## Descriptives tables - Create tables from scratch


In R, it is relatively easy to construct any table you can think of by manipulating objects. To construct a table from scratch, we will use two functions:

 * `aggregate()` - Similar to collapse in Stata, it can compute statistics of a variable based on the values of other variable
 * `reshape()` - Reshapes data sets from long to wide and vice-versa

## Descriptives tables - Create tables from scratch

`aggregate(X, by, FUN)`:

 * __x__: a data frame or column
 * __by__: a list of grouping variables
 * __FUN__: a function to compute statistics

## Descriptives tables - Create tables from scratch

\begin{block}{Exercise 5}
 Use the \texttt{aggregate} function to create a data frame called \texttt{year\_inc\_tab} with the mean of the total income per year and treatment status. The syntax of the \texttt{aggregate} function is very similar to that of the \texttt{collapse} function in Stata.
\end{block}

```{r, eval = T, results = "hide"}
# Aggregate income by year and treatment status
year_inc_tab <-
  aggregate(x = lwh$income_total_win, # data.frame
            by = list(year = lwh$year, # list
                      treatment = lwh$treatment_hh),
            FUN = mean) # function
```

Note that the ``income_total_win`` variable is now named ``x`` in the ``income`` data frame

## Descriptives tables - Create tables from scratch
\scriptsize

```{r}
print(year_inc_tab)

```

## Descriptives tables - Create tables from scratch

`reshape(data, varying, idvar, timevar, direction)`:

 * __data__: a data frame
 * __idvar__: the variables that identify the group in the wide data set
 * __timevar__: the variable in long format that differentiates multiple records from the same group or individual

## Descriptives tables - Create tables from scratch

\begin{block}{Exercise 6}
 Use the \texttt{reshape} function to make the \texttt{year\_inc\_tab} data frame wide per treatment status.
\end{block}


```{r, eval = T, results = "hide"}
# Aggregate income by year and treatment status
year_inc_tab <- reshape(year_inc_tab,
                        idvar = "treatment",
                        timevar = "year",
                        direction = "wide")
```

For comparison, here's how you'd do it in Stata:

``reshape wide x, i(year) j(treatment_hh)``


## Descriptives tables - Create tables from scratch
\scriptsize

```{r}
print(year_inc_tab)

```

## Descriptives tables - Create tables from scratch


With a data frame as input, `stargazer` by default tries to summarize it. So, to export this table we must specify one additional argument: `summary = F`.

### Exercise 7
Print the ``year_inc_tab`` table you created in exercise 6 using ``stargazer``. If you want, you can also save it using the ``out`` option.


## Descriptives tables - Create tables from scratch

\scriptsize

```{r, results='asis'}
# Label variables
column_lab <- c("Treatment status", "2012", "2013", "2014", "2016", "2018")

# Create table
stargazer(year_inc_tab,
          summary = F,
          # Some extra formatting:
          covariate.labels = column_lab,
          title = "Total income by treatment status and year",
          header = F,
          digits = 1,
          rownames = F)
```


# Export tables to Excel

## Export tables to Excel


To export a table to excel we'll use the `write.table()` function. It takes a data frame object as input and saves it as a ``.csv`` file

`write.table()` is the most basic function, but there are many other functions that allow you to export formatted tables to Microsoft Excel, Word or PowerPoint. Here are some examples:

  * ReporteRs
  * Flextable
  * r2excel (only available in GitHub).

## Export tables to Excel

`write.table(x, file = "", sep = " ", row.names = TRUE)`

 * __x__:	the object to be written
 * __file__: where to save the table, i.e., the file path including the file name
 * __sep__: the field separator of the csv, Excel's default is comma
 * __row.names__: either a logical value indicating whether the row names of x are to be written along with x, or a character vector of row names to be written
 * __row.names__: same as `row.names` for columns

## Export tables to Excel

### Exercise 8

Use the `write.table()` function to save the `year_inc_tab` you table created in Exercise 6 into a csv file.

1. Set `x` arugment as *year_inc_tab*.
2. Set `row.names` as `FALSE`
3. Set `col.names` as a vector of labels
4. Set `file` as the folder path to your output folder plus a name for a file plus ".csv"
5. Set `sep` as `","`.

 Tips:

  * Make sure to save it in the *Raw Ouput* folder. You can you the function ``file.path`` to do it
  * Use the help function to check syntax if needed

## Export tables to Excel
\scriptsize

```{r, include = T, eval = T}
write.table(year_inc_tab,
            sep = ",",
            row.names = F,
            col.names = c("Treatment status",
                          "2012", "2013", "2014", "2016", "2018"),
            file = file.path(rawOutput, "year_inc_tab.csv"))

```

\begin{figure}[H]
  \centering
  \includegraphics[scale=.8]{img/year_inc_tab_csv.png}
\end{figure}


# Export regression table

## Export regression table

This is a session on *descriptive* analysis, so regressions are beyond its scope. But since you'll probably ask, here's how you run a regression and how you export a very simple regression table to \LaTeX\ using ``stargazer``:

```{r}
# Run a Regression
reg1 <- lm(expend_food_yearly ~
           income_total_win + num_dependents,
           data = lwh)

```

## Export regression table
\scriptsize


```{r, eval = F}
# Export a regression table

depvar_label <- "Yearly food expenditure (winsorized)"
covar_labels <- c("Total income (winsorized)",
                  "Number of dependents")

stargazer(reg1,
          title = "Regression table",
          dep.var.labels = depvar_label,
          covar_labels = covar_labels,
          digits = 2,
          header = F)

```



## Export regression table
\scriptsize

```{r, results ='asis', echo = F}
# Export a regression table
depvar_label <- "Yearly food expenditure (winsorized)"
covar_labels <- c("Total income (winsorized)", "Number of dependents")

stargazer(reg1,
          title = "Regression table",
          dep.var.labels = depvar_label,
          covariate.labels = covar_labels,
          digits = 2,
          header = F)
```

## Export regression table
\scriptsize

```{r, eval = F}

 
# Regression 1 
reg1 <- lm(expend_food_yearly ~
           income_total_win + num_dependents,
           data = lwh)

# Reg with year FE
reg2 <- lm(expend_food_yearly ~
           income_total_win + num_dependents + factor(year),
           data = lwh)

# Reg with year and site FE
reg3 <- lm(expend_food_yearly ~
           income_total_win + num_dependents + factor(year) + factor(site_code),
           data = lwh)

```

## Export regression table
\scriptsize

```{r, eval = F}
# Labels
depvar_label <- "Yearly food expenditure (winsorized)"
covar_labels <- c("Total income (winsorized)", "Number of dependents")

# Table
stargazer(reg1,
          reg2,
          reg3,
          font.size = "tiny",
          title = "Regression table",
          keep = c("ncome_total_win","num_dependents"),
          dep.var.labels = depvar_label,
          covariate.labels = covar_labels,
          add.lines = list(c("Year FE", "No", "Yes", "Yes"),
                           c("Site FE", "No", "No", "Yes")),
          omit.stat = c("ser"),
          digits = 2,
          header = F)
```

## Export regression table
\scriptsize

```{r, results ='asis', echo = F}


# Run a Regression
reg1 <- lm(expend_food_yearly ~
           income_total_win + num_dependents,
           data = lwh)

# Run a Regression
reg2 <- lm(expend_food_yearly ~
           income_total_win + num_dependents + factor(year),
           data = lwh)

# Run a Regression
reg3 <- lm(expend_food_yearly ~
           income_total_win + num_dependents + factor(year) + factor(site_code),
           data = lwh)

depvar_label <- "Yearly food expenditure (winsorized)"
covar_labels <- c("Total income (winsorized)", "Number of dependents")

stargazer(reg1,
          reg2,
          reg3,
          font.size = "tiny",
          title = "Regression table",
          keep = c("ncome_total_win","num_dependents"),
          dep.var.labels = depvar_label,
          covariate.labels = covar_labels,
          add.lines = list(c("Year FE", "No", "Yes", "Yes"),
                           c("Site FE", "No", "No", "Yes")),
          digits = 2,
          omit.stat = c("ser"),
          header = F)
```

# References and recommendations

## References and recommendations


  * Johns Hopkins Exploratory Data Analysis at Coursera:
  https://www.coursera.org/learn/exploratory-data-analysis

  * Udacity's Data Analysis with R:
  https://www.udacity.com/course/data-analysis-with-r--ud651

  * Jake Russ stargazer cheat sheet:
  https://www.jakeruss.com/cheatsheets/stargazer/

## References and recommendations

  Since we talked about \LaTeX\ so much...

  * DIME \LaTeX\ templates and trainings:
  https://github.com/worldbank/DIME-LaTeX-Templates

  * All you need to know about \LaTeX\:
  https://en.wikibooks.org/wiki/LaTeX
