---
title: "Data Visualization"
subtitle: "R for Stata Users"
date: "June 2019"
author: "Luiza Andrade, Leonardo Viotti & Rob Marty "
output:
  beamer_presentation:
    #theme: "Pittsburgh"
    theme: "Madrid"
    colortheme: "whale"
    fonttheme: "default"
    #toc: true
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F, eval= T, include=FALSE}
# Packages
library(ggplot2)

# File paths

if (Sys.getenv("USERNAME") == "luiza"){
  projectFolder  <- "C:/Users/luiza/Documents/GitHub/dime-r-training"
  
}

if (Sys.getenv("USERNAME") == "WB501238"){
  projectFolder  <- "C:/Users/WB501238/Documents/GitHub/dime-r-training"
  
}

if (Sys.getenv("USERNAME") == "Leonardo"){
  projectFolder  <- "C:/Users/Leonardo/Documents/GitHub/dime-r-training"

}

if (Sys.getenv("USERNAME") == "wb519128"){
  projectFolder <- file.path("C:/Users/WB519128/Documents/GitHub/dime-r-training")
}

# File paths
  dataWorkFolder    <- file.path(projectFolder,"DataWork")

  Data              <- file.path(dataWorkFolder,"DataSets")
  finalData         <- file.path(Data,"Final")
  rawOutput         <- file.path(dataWorkFolder,"Output","Raw")

# Load CSV data

whr <- read.csv(file.path(finalData,"whr_panel.csv"), 
                header = T)


```


# Introduction

## Introduction

In this session, you'll learn how to use R to produce insightful, meaningful and (hopefully) nice-looking graphs. In particular, you'll use a package called ``ggplot2``.

Similarly to the previous session, you can find some references at the end of this presentation that include a more comprehensive discussions on data visualization.

## Introduction

Before we start, let's make sure we're all set:

\begin{enumerate}

  \item Make sure you the packages \texttt{ggplot2} and \texttt{plotly} are installed and loaded.

  \item Load the \texttt{whr\_panel.csv} data set.

\end{enumerate}

## Introduction
\scriptsize

```{r, eval= F}
  # Install packages
  install.packages("plotly", # We already installed ggplot2
                   dependencies = TRUE)
  
  # Load packages
  library(ggplot2)
  library(plotly)

  #### Load CSV data

  # Replace with where your data is
  FolderPath <- file.path("YOUR/FOLDER/PATH")

  whr <- read.csv(file.path(FolderPath,"whr_panel.csv"), 
                  header = T,
                  stringsAsFactors = F)  

```

## Introduction

In our workflow there are usually two distinct uses for plots:

  \begin{enumerate}
    \item \textbf{Exploratory analysis:} Quickly visualize your data in an insightful way
      \begin{itemize}
        \item We'll do this using R's base plot, that allows you to quickly create some basic plots
      \end{itemize}
    \item \textbf{Publication/Reporting:} Make pretty graphs for a presentation, a project report, or to just show your boss something other than the laziest graph you could come up with
      \begin{itemize}
        \item We'll do this using \texttt{ggplot}, a flexible and powerful tool to plot beautiful graphs
        \item \texttt{ggplot}'s syntax is more complicated, but it's easier to make your graphs look good with it
      \end{itemize}
  \end{enumerate}


# Exploratory graphs - Base plot

## Exploratory graphs - Base plot

  For exploratory plots, we're going to use Base plot. It is easy to use and can produce useful graphs with very few lines of code.

## Exploratory graphs - Base plot

### Exercise 1:
Plot the ``whr_simp`` data set you constructed in the previous session using the ``plot()`` function.

 1. If you don't have the code from the previous session. Here it is:

```{r, eval = T}
# Vector with covariates to be kept
covariates <- c("happy_score",
                "gdp_pc",
                "freedom",
                "trust_gov_corr")

# subset whr
whr_simp <- whr[, covariates]

```
 2. Now, pass the name of the data set as the only argument for the plot function

## Exploratory graphs - Base plot

\scriptsize

```{r, out.width = "85%"}
plot(whr_simp)
```

## Exploratory graphs - Base plot

\begin{block}{Exercise 2}
  \begin{enumerate}
    \item Create a histogram of variable \texttt{happy\_score} in data frame \texttt{whr\_simp}. Use the \texttt{hist()} function for that.
    \item Try to set the color as \texttt{"green"}. Use the help if you are not sure how to do it.
  \end{enumerate}
\end{block}


## Exploratory graphs - Base plot
\scriptsize
```{r, eval = T, echo = T, out.width = "85%"}
hist(whr_simp$happy_score, col = "green")
```

## Exploratory graphs - Base plot

### Exercise 3
Use the ``boxplot()`` function to see a description of the happiness score by region.

- TIP: The syntax for this function is ``variable_to_be_plotted`` ~ ``variable_to_plot_over``.


## Exploratory graphs - Base plot


Since these region labels are too long, I'm using an extra argument so we can actually read the x axis.

```{r, eval = F}
boxplot(whr$happy_score ~ whr$region, las = 2)
```



## Exploratory graphs - Base plot
\tiny

```{r, echo = F, out.width = "95%"}
par(mar=c(14,2,2,2))
boxplot(whr$happy_score ~ whr$region, las = 2)
```

# ``ggplot`` - Introduction

## ``ggplot`` - Introduction


The ``ggplot2`` package is an implementation of the theoretical framework for the construction of quantitative graphs developed in The Grammar of Graphics, a book by Leland Wilkinson. 

It is a powerful and easy to use tool (once you understand its logic) that produces complex and multifaceted plots.

## ``ggplot`` - Introduction

There are a few reasons why we use ggplot:

 * It is generally easier to customize and tailor plots to your needs.
 * It works great with almost any kind of plot, including maps, with little variation in syntax.
 * It looks good.

## ``ggplot`` - Grammar

Here is a list of ``ggplot2`` most commonly used "grammar" elements:


\begin{table}[]
\centering
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|m{2cm}| m{8cm}|}
  \hline
  \rowcolor{myColor}
  \multicolumn{1}{|c|}{\textcolor{white}{Element}}   &      \multicolumn{1}{|c|}{\textcolor{white}{Description}} \\
  Data       & The dataset(s) being used.  \\ \hline
  Aesthetics & How your data is maped. For example what goes in the X and Y axis, size, shape and color. \\ \hline
  Geometries & The visual elements used to represent the data (e.g. lines, dots, bars, etc.) \\ \hline
  Themes     & All non-data formating. \\
  \hline
  \end{tabular}
\end{table}




## ``ggplot`` - Introduction


### Exercise 4

First, let's create a very simple plot with the happiness score on the Y axis and freedom on the X axis.


 1. Use the ``ggplot`` function to store your plot in an object called ``p1_happyfree``
    - Since it's the first one, here's the code:

```{r, eval=F}
p1_happyfree <- 
  ggplot(data = whr,
         aes(y = happy_score,
             x = freedom))
  
```

 2. Now, try to print your plot. What happens?
    - You can do it by using the ``print()`` function or just typing the object name in the console

## ``ggplot`` - Introduction

We mapped our variables, but without a geometry ``ggplot`` doesn't know how to represent the data.

### Let's add a geometry
```{r, eval=TRUE}
p1_happyfree <- 
  ggplot(whr, 
         aes(y = happy_score,
             x = freedom)) +
  geom_point()
```


## ``ggplot`` - Introduction
\scriptsize

```{r, echo = T, warning = F, out.width = "88%"}
print(p1_happyfree)
```

# ``ggplot`` - Data structure

## ``ggplot`` - Data structure

As almost everything in R, ``ggplot`` is very flexible and can be used in different ways. It is easier to work, however, if your data is in a specific format.

  * Data has to be a data frame.

  * Long format, especially categorical values.
    + It's better to have one variable with all the categories than a set of dummies.
  * Labelled factors.

This is very similar to Stata, with the advantage of not needing to
preserve and restore anything.

## ``ggplot`` - Data structure

Here's an example of how an aggregated (collapsed) data set should look like:


```{r, include = T, echo = F, warning=FALSE}
happy_table <-
  aggregate(happy_score ~ year + region,
            data = whr,
            FUN = mean)
happy_table[1:11,]

```

## ``ggplot`` - Data structure

* Since ``ggplot`` usually works better with data in a specific format, there are a few functions particularly useful to manipulate data
* Here, we'll use the ``aggregate()`` function
* ``aggregate()`` works similarly to ``collapse`` in Stata

## ``ggplot`` - Data structure

To create a line plot with ``ggplot``, we will create a new data set containing the average of happiness score by year.
```{r, eval= T}
# Aggregate annual income by year
annualHappy<-
  aggregate(happy_score ~ year,
            data = whr,
            FUN = mean)

head(annualHappy)
```

## ``ggplot`` - Data structure

### Exercise 5
 1. Use the ``ggplot()`` function with the ``annualHappy`` as the ``data`` argument, ``happy_score`` as ``y`` and ``year`` as ``x`` in ``aes()``
 2. This time, add ``geom_line()``


## ``ggplot`` - Data structure

```{r,  eval= T}

# Plot code with arguments in geom
p2_happyyear <-
  ggplot(data = annualHappy, 
         aes(y = happy_score,
             x = year)) +
  geom_line()

```

## ``ggplot`` - Data structure
\scriptsize

```{r, echo = T , eval= T, out.width = "88%"}

print(p2_happyyear)



```


## ``ggplot`` - Data structure

Ok, it worked, but that was a really boring plot. Let's do a better one.

### Exercise 6: Part 1

Use the ``aggregate()`` function to average happiness score, now, by two variables: `year` and `region`.

```{r, eval = F}
annualHappy_reg <-
  aggregate(happy_score ~ year + region,
            data = whr,
            FUN = mean)
```

## ``ggplot`` - Data structure
\scriptsize

```{r}
# Aggregate data set
annualHappy_reg <-
  aggregate(happy_score ~ year + region,
            data = whr,
            FUN = mean)

annualHappy_reg[1:10,]

```

## ``ggplot`` - Data structure

### Exercise 6: Part 2
 Now, use `ggplot` as in the previous exercise to plot the aggregated data frame.
 
 1. This time set the ``color`` and `group` arguments in the ``aes()`` function as the treatment variable.
 2. Finally, add both line and point ``geoms``.

## ``ggplot`` - Data structure
\scriptsize

```{r,  eval= F}
# Aggregate data set
annualHappy_reg <-
  aggregate(happy_score ~ year + region,
            data = whr,
            FUN = mean)



# Plot aggregated data
p3_happyreg <-
  ggplot(data = annualHappy_reg, 
         aes(y = happy_score,
             x = year, 
             color = region, 
             group = region)) +
    geom_line() +
    geom_point()

print(p3_happyreg)
```

## ``ggplot`` - Data structure
\scriptsize

```{r, echo = F , eval= T, warning=FALSE, out.width = "88%"}

annualHappy_reg <-
  aggregate(happy_score ~ year + region,
            data = whr,
            FUN = mean)

ggplot(annualHappy_reg, 
       aes(y = happy_score,
           x = year, 
           color = region, 
           group = region)) +
  geom_line() +
  geom_point()


```


## ``ggplot`` - Data structure

You can also use `ggplot` to combine different data frames in the same plot.

 * Everything inside the `ggplot()` function will be shared by all geometries added (e.g. the line and point geometries added to the previous exercise)
 
 * But you can also pass different arguments to different geometries.
 
In the following example, to showcase just 3 regions, instead of restricting the data to only the interest regions, we are adding 3 separeate line geometries.


## ``ggplot`` - Data structure
\tiny

```{r, echo = T , eval= T}
# Separate datasets
sea  <- annualHappy_reg[annualHappy_reg$region == "Southeastern Asia", ]
mena <- annualHappy_reg[annualHappy_reg$region == "Middle East and Northern Africa", ]
cee  <- annualHappy_reg[annualHappy_reg$region == "Central and Eastern Europe", ]

# plot with 3 line geometries 
p4_happy3reg <-
  ggplot(mapping = aes(y = happy_score,
                       x = year, 
                       color = region, 
                       group = region)) +
    geom_line(data = sea) +  # Southeastern Asia
    geom_line(data = mena) + # Middle East and Northern Africa
    geom_line(data = cee) # Central and Eastern Europe
```


## ``ggplot`` and ``aggregate()``

\scriptsize

```{r, eval= T, warning=FALSE, out.width = "88%"}
print(p4_happy3reg)
```


# ``ggplot`` - Aesthetics

## ``ggplot`` - Aesthetics

In ``ggplot`` , the aesthetic element defines how data is represented in the aesthetics (e.g. color, shape and size) of geometric objects (e.g. points, lines and bars).

### Exercise 7: Part 1

 1. Create a data frame containing only 2017 observations of `whr`.

  - TIP: This time we want to keep only observations (lines) that meet a certain condition. Use the left argument of the brackets operators. Like this: 
```{r, eval = F}
    whr[CONDITION,]
```

 2. Use ``ggplot()`` to scatter happiness score and freedom.

## ``ggplot`` - Aesthetics
\scriptsize

```{r,  eval= T}
# Subset whr to 2017
whr17 <- whr[whr$year == 2017, ]

# Plot
p5_happyfree17 <-
  ggplot(whr17, aes(x = freedom,
                    y = happy_score)) +
      geom_point()

```

## ``ggplot`` - Aesthetics
\scriptsize

```{r, echo = T , eval= T , warning=FALSE, out.width = "88%"}
print(p5_happyfree17)


```


## `ggplot` - Aesthetics

Now add the region to `aes()`.

### Exercise 7: Part 2

 1. Use the `region` variable as color in ``aes()``.
    - TIP: this is is a categorical variable, but it is stored in string format.  Use the ``factor()`` function. Like this:
```{r, echo = T , eval= F}
    color = factor(region)
```


## ``ggplot`` - Aesthetics
\scriptsize

```{r,  eval= T}
# Plot code
p6_happyfree17reg <- 
  ggplot(whr17, 
         aes(x = freedom,
             y = happy_score)) +
  geom_point(aes(color = factor(region)))


```

## ``ggplot`` - Aesthetics
\scriptsize

```{r, echo = F , eval= T, warning=FALSE, out.width = "88%"}

print(p6_happyfree17reg)

```

## ``ggplot`` - Aesthetics

Now we will combine different arguments of ``aes()``

### Exercise 7: Part 3

 1. Add `gdp_pc` to the ``size`` argument of ``aes()`` in your last graph.

## ``ggplot`` - Aesthetics
\scriptsize

```{r,  eval= T}
# Plot code
p6_happyfree17reg <-
  ggplot(whr17, 
         aes(x = freedom,
             y = happy_score)) +
  geom_point(aes(color = factor(region),
                 size = gdp_pc))

```

## ``ggplot`` - Aesthetics
\scriptsize

```{r, echo = T , eval= T, warning=FALSE, out.width = "88%"}

print(p6_happyfree17reg)

```

## ``ggplot`` - Aesthetics

* Because we used a factor variable as the color in the aesthetic, it displayed different categories clearly
* If we had used a numeric variable instead, it would have created a gradient
* You'll not create this for time's sake, but here's how it would look:

```{r,  eval= F}
# Plot code
ggplot(whr17, aes(x = freedom,
                  y = happy_score,
                  color = dystopia_res,
                  size = gdp_pc)) +
    geom_point()

```

## ``ggplot`` - Aesthetics
\scriptsize

```{r, echo = F , eval= T, warning=FALSE, out.width = "88%"}

ggplot(whr17, aes(x = freedom,
                  y = happy_score,
                  color = dystopia_res,
                  size = gdp_pc)) +
    geom_point()

```

# ``ggplot`` - Titles and labels

## ``ggplot`` - Titles and labels

Lastly, we can add titles, axis labels and format legends. This is done by adding additional layers to the plot.

To add a title and legend to the X and Y axis, we use the functions listed below. They all take a string scalar as argument.

 * ``ggtitle()`` - adds a title to the plot
 * ``xlab()`` - adds a label to X axis
 * ``ylab()`` - adds a label to Y axis

## ``ggplot`` - Titles and labels

To add legend titles, we will change the formatting of the aesthetic. We will use `labs()` function specifying wich aesthetic elements we will modify.

 ``labs(color = "Color lab", fill = "Fill lab", ...) ``

## ``ggplot`` - Titles and labels

#### Exercise 8

 1. Copy the code for the graph produced in the previous exercise.
 2. Use the ``+`` symbol to add the layer.
 3. ``ggtitle()``, ``xlab()`` and ``ylab()`` take simple strings as inputs.
 4. For labelling the legends use `color` and `size` arguments of the `labs()` function.

## ``ggplot`` - Titles and labels
\scriptsize

```{r, eval= T}
# A properly labelled plot
p6_happyfree17reg <-
  p6_happyfree17reg +
      ggtitle("My pretty plot") +
      xlab("Freedom") +
      ylab("Happiness") +
      labs(color = "World Region",
           size = "Score of GDP per capita")
```


## ``ggplot`` - Titles and labels
\scriptsize

```{r, echo = T , eval= T,  warning=FALSE, out.width = "88%"}
p6_happyfree17reg
```


# Saving a plot

## Saving a plot

Let's save the previous plot

#### Exercise 9: We will use the `ggsave()` function to save a .png image
  1. "Add" the `ggsave()` function to the `p6_happyfree17reg` object you created.
  2. Set the ``filename`` argument as your outputs path with the file name (including ".png")
  4. Specify the `width` and `height` arguments. The default unit is inches, but you can use the `units` arguments to specify diferently.

#### TIP:
If you don't specify the `width` and `height`, `ggsave()` will use the size of Rstudio's graphics device window.



## Saving a plot
\scriptsize

```{r, eval= F}

Output <- "YOUR/PATH/HERE"

p6_happyfree17reg + 
  ggsave(filename = file.path(Output, "myPrettyGraph.png"), 
         width = 9, 
         height = 6)

```




# Interactive graphs


## Interactive graphs

There are several packages to create interactive or dynamic data vizualizations with R. Here are a few:

 * `leaflet` - R integration tp one of the most popular open-source libraries for interactive maps.
 * `highcharter` - cool interactive graphs.
 * `plotly` - interactive graphs with integration to ggplot.
 * `gganimate` - ggplot GIFs.
 * `DT` - Interactive table

These are generally, html widgets that can be incorporated in to an html document and websites.

## Interactive graphs

Now we'll use the `ggplotly()` function from the `plotly` package to create an interactive graph!

### Exercise 9
  1. Load the ``plotly`` package
  2. Pass that object with the last plot you created to the ``ggplotly()`` function

## Interactive graphs
\scriptsize

```{r,  eval= F}
# Load package
library(plotly)

 # Use ggplotly to create an interactive plot
 ggplotly(p6_happyfree17reg)
```

## Interactive graphs

(Sorry, this is a .pdf presentation so this is just a screenshot.)

\begin{figure}
\centering
  \includegraphics[width=12cm,height=7.7cm]{img/ggplotly1.png}
\end{figure}



## Homework - Interactive graphs

We can add pop-up information by adding another argument to the `aes()` function. 

### Customize your ggplotly plot
  1. Add a `text` argument to the`geom_point` `aes()` function and assign it to the `country` variable. This argument won't be recognized by `ggplot()`, but will be by `ggplotly()`.
  2. Use themes to change the background color. You can find a list of default themes in https://ggplot2.tidyverse.org/reference/ggtheme.html
  
  - TIP: type `?ggthemes::` on the console for more options
  
  3. Use the `tooltip` argument of the `ggplotly` function to customize what is shown when hovering

## Interactive graphs

Here's what we've done:

\begin{figure}
\centering
  \includegraphics[width=11cm,height=7cm]{img/ggplotly2.png}
\end{figure}


## Interactive graphs

(Sorry, this is a .pdf presentation so this is just a screenshot.)

\begin{figure}
\centering
  \includegraphics[width=11cm,height=7cm]{img/ggplotly2.png}
\end{figure}

# References and recommendations

## References and recommendations

Websites:

  * Interactive stuff :
  http://www.htmlwidgets.org/

  * The R Graph Gallery:
  https://www.r-graph-gallery.com/

  * Gpplot official site:
  http://ggplot2.tidyverse.org/


Online courses:

  * Johns Hopkins Exploratory Data Analysis at Coursera:
  https://www.coursera.org/learn/exploratory-data-analysis

Books:

  * The grammar of graphics by Leland Wilkinson.

  * Beautiful Evidence by Edward Tufte.

  * R Graphics cook book by Winston Chang

  * R for Data Science by Hadley Wickham andGarrett Grolemund

# Appendix

## Tpyes of plots and themes

### Choose your plot 
 - Based on your data and what you want to show choose a type of plot, `geom_density()`, `geom_boxplot`, `geom_point`, etc.
 - You can find a list of plots in https://ggplot2.tidyverse.org/reference/

Themes:
\scriptsize
```{r,  eval= F}
# Install and load package
install.packages("ggthemes") #as the scheme in Stata
library(ggthemes)

 # Use theme with ggplot
ggplot(whr17, aes(x = freedom,
                  y = happy_score,
                  color = dystopia_res,
                  size = gdp_pc)) +
    geom_point()+
    theme_light()
```
## Color palettes

### Customize the color of your plot
There are several color palettes in R
- RColorBrewer https://rdrr.io/cran/RColorBrewer/man/ColorBrewer.html 
- Viridis https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
- Wesanderson https://rdrr.io/cran/wesanderson/src/R/colors.R

```{r,  eval= F}
# Load package
library(RColorBrewer)
par(mar=c(3,4,2,2))
#Print palettes
display.brewer.all()

#Load package
library(viridis)

# Load package
library(wesanderson)
#Print palettes 
names(wes_palettes)

```

## R common graphic devices

R has several built-in graphic devices. A graphic device is the medium where you visually represent a plot. Here are the most come:

  * The standard screen device or the plot pane in RStudio.
  * PDF file (file device).
  * PNG (file device).
  * JPEG (file device).

## Differed ways of saving a plot

There are a couple of ways of saving plots in R:

 * With ggplot you can use the ``ggsave()`` function that has a simple syntax and can do most of the dirty work for you.

 * Choosing an specific graphics device (other than the default window). pdf(), png(), jpeg() and bmp() are functions that call graphic devices that save your plots in these formats. There are a few others built-in and you can check them by typing *?Devices*, and even more in CRAN.


## Graphic systems

R has four main graphic systems:

  * Base plot - most basic plotting system.
  * Grid graphics - is a low-level graphics system mainly used by package developers.
  * Lattice graphics - easy way to plot high dimensional data or many graphs at the same time.
  * Ggplot2 - a flexible and powerful tool to plot beautiful graphs with intuitive syntax

